% ============================================================================
\section{Understanding Transformers and Large Language Models}
\label{sec:transformers-llms}
% ============================================================================

The previous sections developed four paradigms for replacing
floating-point arithmetic with bitwise operations, and showed how they
combine into a hybrid architecture for MNIST. Before we can honestly
evaluate whether any of this scales to large language models (LLMs),
we need to understand what LLMs actually \emph{are}---what makes them
work, and which of their components are essential versus replaceable.

This section builds that understanding from the ground up, using the
same intuition-first approach as the rest of this article.

% ----------------------------------------------------------------------------
\subsection{From Pixels to Words: Why Language Is Harder}
\label{sec:pixels-vs-words}
% ----------------------------------------------------------------------------

MNIST is a \emph{fixed-size classification} problem: 784 pixels in,
one of ten labels out. Every input has the same shape, and every pixel
has a clear spatial relationship to its neighbours.

Language is fundamentally different:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Variable length.} A sentence can be 5~words or
    5{,}000~words. The network must handle both.
  \item \textbf{No fixed grid.} Words are not arranged in a 2D grid
    like pixels. Their meaning depends on \emph{order} and
    \emph{context}---``the dog bit the man'' and ``the man bit the
    dog'' use the same words but mean different things.
  \item \textbf{Long-range dependencies.} In ``The cat that the dog
    chased ran away,'' the verb ``ran'' relates to ``cat,'' not the
    nearby ``dog.'' The network must connect words across arbitrary
    distances.
  \item \textbf{Generation, not classification.} An LLM does not pick
    one of ten labels. It generates text one word at a time, choosing
    from a vocabulary of 30{,}000--100{,}000 possible next words at
    each step.
\end{itemize}

\begin{keyinsight}
The transition from MNIST to language is not just ``bigger data.'' It
requires fundamentally different architectural primitives: a way to
represent words as numbers, a mechanism to relate words to each other
regardless of distance, and a method for generating sequences one
element at a time.
\end{keyinsight}

% ----------------------------------------------------------------------------
\subsection{Tokenisation: Turning Text into Numbers}
\label{sec:tokenisation}
% ----------------------------------------------------------------------------

A neural network cannot process letters or words directly---it needs
numbers. The first step in any language model is \concept{tokenisation}:
splitting text into small pieces and assigning each piece an integer ID.

Modern LLMs do not split on whole words. Instead, they use
\concept{subword tokenisation} (algorithms like Byte-Pair Encoding or
SentencePiece), which breaks text into frequently occurring fragments:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  tok/.style={draw, rounded corners=2pt, minimum height=0.7cm,
    font=\small\ttfamily, inner sep=4pt, fill=#1!10, draw=#1!60},
  arr/.style={-{Stealth[length=4pt]}, thick, bitgray}
]
  % Original text
  \node[font=\small\sffamily] at (-1.5, 1.5) {Input text:};
  \node[font=\small\ttfamily] at (4, 1.5)
    {``Understanding transformers is fascinating''};

  % Tokens
  \node[font=\small\sffamily] at (-1.5, 0) {Tokens:};
  \node[tok=bitblue]   (t1) at (1.5, 0) {Under};
  \node[tok=bitgreen]  (t2) at (3.3, 0) {stand};
  \node[tok=bitorange] (t3) at (5.0, 0) {ing};
  \node[tok=bitpurple] (t4) at (7.0, 0) {\_transform};
  \node[tok=bitblue]   (t5) at (9.3, 0) {ers};
  \node[tok=bitgreen]  (t6) at (11.0, 0) {\_is};
  \node[tok=bitorange] (t7) at (13.0, 0) {\_fasc};
  \node[tok=bitpurple] (t8) at (14.8, 0) {inating};

  % Token IDs
  \node[font=\small\sffamily] at (-1.5, -1.2) {Token IDs:};
  \node[font=\small\ttfamily] at (1.5, -1.2) {8803};
  \node[font=\small\ttfamily] at (3.3, -1.2) {1514};
  \node[font=\small\ttfamily] at (5.0, -1.2) {278};
  \node[font=\small\ttfamily] at (7.0, -1.2) {13389};
  \node[font=\small\ttfamily] at (9.3, -1.2) {414};
  \node[font=\small\ttfamily] at (11.0, -1.2) {338};
  \node[font=\small\ttfamily] at (13.0, -1.2) {19239};
  \node[font=\small\ttfamily] at (14.8, -1.2) {15536};
\end{tikzpicture}
\caption{Subword tokenisation splits text into frequently occurring
  fragments, each mapped to an integer ID. Common words become single
  tokens; rare words are split into subwords. The underscore
  (\_) represents a space preceding the token.}
\label{fig:tokenisation}
\end{figure}

\begin{analogy}
Tokenisation is like Morse code for neural networks. Just as Morse
represents each letter as a sequence of dots and dashes, tokenisation
represents each text fragment as a number. The difference is that the
``alphabet'' is learned from data: common fragments (``the'', ``ing'',
``tion'') get their own codes, while rare fragments are spelled out
from smaller pieces. A typical LLM vocabulary has 30{,}000--100{,}000
tokens.
\end{analogy}

The key property is that \emph{any} text can be represented as a
sequence of token IDs. The model's input is always a sequence of
integers: $[t_1, t_2, \ldots, t_n]$ where each $t_i \in
\{0, 1, \ldots, V{-}1\}$ and $V$ is the vocabulary size.

\subsubsection{How the Vocabulary Is Constructed: BPE}

The vocabulary is \emph{not} learned during neural network
training---it is constructed \textbf{beforehand} as a separate
preprocessing step, then fixed for the lifetime of the model.
The most common algorithm is \concept{Byte-Pair Encoding} (BPE),
which works as follows:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Start with bytes.} The initial vocabulary consists of
    all 256 individual bytes (characters). Every possible text can
    already be represented, just very inefficiently---one token per
    character.

  \item \textbf{Count pairs.} Scan the entire training corpus and
    count how often each pair of adjacent tokens appears. For example,
    in English text the pair (``t'', ``h'') appears far more often
    than (``q'', ``z'').

  \item \textbf{Merge the most frequent pair.} Create a new token
    by combining the most common pair. For instance, ``t'' + ``h''
    $\to$ ``th'' becomes token number~257.

  \item \textbf{Repeat.} Go back to step~2 with the updated
    vocabulary. Now the algorithm might find that (``th'', ``e'') is
    the most frequent pair, so it creates ``the'' as token~258.
    Continue until the vocabulary reaches the desired size
    (e.g.\ 32{,}000 tokens).
\end{enumerate}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  steplbl/.style={font=\scriptsize\sffamily, anchor=west},
  tok/.style={draw, rounded corners=1pt, minimum height=0.45cm,
    inner sep=3pt, font=\scriptsize\ttfamily, fill=#1!15},
  arr/.style={->, thick, bitgray}
]
  % Step 0
  \node[steplbl, font=\scriptsize\sffamily\bfseries] at (-1, 0) {Start:};
  \foreach \i/\c in {1/l, 2/o, 3/w, 4/e, 5/r, 6/ , 7/l, 8/o,
    9/w, 10/e, 11/s, 12/t} {
    \node[tok=bitgray] at ({\i*0.55+0.5}, 0) {\c};
  }
  \node[steplbl, gray] at (8, 0) {(256 tokens)};

  % Step 1
  \node[steplbl, font=\scriptsize\sffamily\bfseries] at (-1, -1)
    {Merge 1:};
  \node[steplbl, bitorange] at (-1, -1.5) {``l''+``o'' $\to$ ``lo''};
  \foreach \i/\c/\col in {1/lo/bitorange, 2/w/bitgray, 3/e/bitgray,
    4/r/bitgray, 5/ /bitgray, 6/lo/bitorange, 7/w/bitgray,
    8/e/bitgray, 9/s/bitgray, 10/t/bitgray} {
    \node[tok=\col] at ({\i*0.6+0.5}, -1) {\c};
  }
  \node[steplbl, gray] at (8, -1) {(257 tokens)};

  % Step 2
  \node[steplbl, font=\scriptsize\sffamily\bfseries] at (-1, -2.5)
    {Merge 2:};
  \node[steplbl, bitblue] at (-1, -3) {``lo''+``w'' $\to$ ``low''};
  \foreach \i/\c/\col in {1/low/bitblue, 2/e/bitgray,
    3/r/bitgray, 4/ /bitgray, 5/low/bitblue,
    6/e/bitgray, 7/s/bitgray, 8/t/bitgray} {
    \node[tok=\col] at ({\i*0.7+0.5}, -2.5) {\c};
  }
  \node[steplbl, gray] at (8, -2.5) {(258 tokens)};

  % Step 3
  \node[steplbl, font=\scriptsize\sffamily\bfseries] at (-1, -4)
    {Merge 3:};
  \node[steplbl, bitgreen] at (-1, -4.5)
    {``low''+``e'' $\to$ ``lowe''};
  \foreach \i/\c/\col in {1/lowe/bitgreen,
    2/r/bitgray, 3/ /bitgray, 4/lowe/bitgreen,
    5/s/bitgray, 6/t/bitgray} {
    \node[tok=\col] at ({\i*0.8+0.5}, -4) {\c};
  }
  \node[steplbl, gray] at (8, -4) {(259 tokens)};

  % Step 4
  \node[steplbl, font=\scriptsize\sffamily\bfseries] at (-1, -5.5)
    {Merge 4:};
  \node[steplbl, bitpurple] at (-1, -6)
    {``lowe''+``r'' $\to$ ``lower''};
  \foreach \i/\c/\col in {1/lower/bitpurple,
    2/ /bitgray, 3/lowe/bitgreen,
    4/s/bitgray, 5/t/bitgray} {
    \node[tok=\col] at ({\i*0.9+0.8}, -5.5) {\c};
  }
  \node[steplbl, gray] at (8, -5.5) {(260 tokens)};

  % Annotation
  \draw[decorate, decoration={brace, amplitude=4pt},
    thick, bitgray]
    (10.5, 0.3) -- (10.5, -5.8);
  \node[font=\scriptsize\sffamily, bitgray, anchor=west,
    text width=3cm, align=left] at (10.8, -2.75)
    {repeat until\\vocabulary has\\32{,}000 tokens};
\end{tikzpicture}
\caption{BPE vocabulary construction on the text ``lower lowest.''
  Starting from individual characters, the algorithm repeatedly merges
  the most frequent adjacent pair. After four merges, ``lower'' has
  become a single token while ``lowest'' is still two tokens
  (``lowe'' + ``st''). Common words get merged first; rare words stay
  split.}
\label{fig:bpe}
\end{figure}

The result is a fixed lookup table mapping token IDs to string
fragments. Several properties make BPE well-suited for LLMs:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Efficiency:} Common words like ``the'' become single
    tokens, so typical English text is compressed to about one token
    per 3--4 characters. This means fewer tokens to process, saving
    computation.
  \item \textbf{Completeness:} Any text can be encoded---even
    misspellings, code, URLs, or languages unseen during vocabulary
    construction---because the byte-level fallback always works.
  \item \textbf{Frequency-adaptive:} Frequent sequences get short
    encodings (single tokens); rare sequences get longer encodings
    (multiple tokens). This is essentially Huffman coding applied to
    text fragments.
\end{itemize}

\begin{engineernote}
The vocabulary is constructed \emph{once} before training begins and
never changes. This means vocabulary design decisions---vocabulary
size, training corpus, algorithm---are locked in permanently. A poor
vocabulary choice cannot be fixed by training longer. This is why
modern models use large vocabularies (100{,}000+ tokens): the extra
embedding parameters are cheap compared to the cost of representing
common words as multiple tokens throughout all transformer layers.
\end{engineernote}

% ----------------------------------------------------------------------------
\subsection{Embeddings: Words as Points in Space}
\label{sec:embeddings}
% ----------------------------------------------------------------------------

A token ID like 13389 is just a number---it tells us nothing about
what the token \emph{means}. The next step is to convert each token
into a \concept{high-dimensional vector}, called an
\concept{embedding}.

An embedding is a list of numbers (typically 768 to 4{,}096 of them)
that represents the token as a point in a high-dimensional space. The
critical property is that tokens with similar meanings end up at
nearby points:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0]
  % Axes
  \draw[-{Stealth}, thick, bitgray] (-0.3, 0) -- (8, 0);
  \draw[-{Stealth}, thick, bitgray] (0, -0.3) -- (0, 6);
  \node[font=\scriptsize\sffamily, bitgray, right] at (8, 0)
    {dimension 1};
  \node[font=\scriptsize\sffamily, bitgray, above] at (0, 6)
    {dimension 2};

  % Animal cluster
  \fill[bitblue!15, rounded corners=10pt]
    (0.5, 3.5) rectangle (3.5, 5.5);
  \node[font=\scriptsize\sffamily, bitblue] at (2, 5.8) {animals};
  \node[circle, fill=bitblue, inner sep=2pt, label={[font=\scriptsize]right:cat}]
    at (1.5, 4.8) {};
  \node[circle, fill=bitblue, inner sep=2pt, label={[font=\scriptsize]right:dog}]
    at (2.2, 4.3) {};
  \node[circle, fill=bitblue, inner sep=2pt, label={[font=\scriptsize]right:horse}]
    at (1.8, 3.9) {};

  % Vehicle cluster
  \fill[bitorange!15, rounded corners=10pt]
    (4.5, 3.5) rectangle (7.5, 5.5);
  \node[font=\scriptsize\sffamily, bitorange] at (6, 5.8) {vehicles};
  \node[circle, fill=bitorange, inner sep=2pt, label={[font=\scriptsize]right:car}]
    at (5.5, 4.8) {};
  \node[circle, fill=bitorange, inner sep=2pt, label={[font=\scriptsize]right:truck}]
    at (6.2, 4.3) {};
  \node[circle, fill=bitorange, inner sep=2pt, label={[font=\scriptsize]right:bus}]
    at (5.8, 3.9) {};

  % Royalty cluster
  \fill[bitpurple!15, rounded corners=10pt]
    (2.5, 0.5) rectangle (6.5, 2.5);
  \node[font=\scriptsize\sffamily, bitpurple] at (4.5, 2.8) {royalty};
  \node[circle, fill=bitpurple, inner sep=2pt, label={[font=\scriptsize]right:king}]
    at (3.5, 1.8) {};
  \node[circle, fill=bitpurple, inner sep=2pt, label={[font=\scriptsize]right:queen}]
    at (4.2, 1.3) {};
  \node[circle, fill=bitpurple, inner sep=2pt, label={[font=\scriptsize]right:prince}]
    at (5.5, 1.5) {};

  % Analogy arrow
  \draw[-{Stealth}, thick, dashed, bitgreen]
    (3.5, 1.8) -- node[font=\scriptsize\sffamily, above, sloped]
    {king $-$ man $+$ woman $\approx$ queen} (4.2, 1.3);

  % Note
  \node[font=\scriptsize\sffamily, bitgray, text width=4cm, align=center]
    at (8, 2.5) {(shown in 2D;\\actual embeddings\\use 768--4096\\dimensions)};
\end{tikzpicture}
\caption{Token embeddings place words in a high-dimensional space
  where distance reflects semantic similarity. Similar words cluster
  together, and vector arithmetic can capture analogies.
  This picture is simplified to 2D; real embeddings use hundreds to
  thousands of dimensions.}
\label{fig:embeddings}
\end{figure}

\begin{engineernote}
An embedding table is conceptually simple: it is a 2D array of shape
$V \times d$, where $V$ is the vocabulary size and $d$ is the
embedding dimension. Looking up a token's embedding is a single
table lookup---no computation, just a memory read of $d$ numbers.
In a model with $V = 50{,}000$ and $d = 4{,}096$, the embedding
table alone uses $50{,}000 \times 4{,}096 \times 2 = 400$\,MB
(in 16-bit precision).
\end{engineernote}

But where do these embeddings come from? Unlike the vocabulary
(which is constructed once before training), the embedding table is
\emph{learned during training} as part of the neural network. Here
is how:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Initialise randomly.} At the start of training, the
    embedding table $\bm{E} \in \mathbb{R}^{V \times d}$ is filled
    with random numbers. Row~4521 (``cat'') is a random vector,
    row~7803 (``dog'') is a different random vector. At this point,
    ``cat'' and ``dog'' are no more similar than ``cat'' and
    ``bicycle.''

  \item \textbf{Forward pass.} A token enters the model. The
    network looks up its row in $\bm{E}$, passes that vector through
    all transformer blocks, and produces a prediction for the next
    token.

  \item \textbf{Compute error.} The prediction is compared to the
    actual next token. The difference is the error (loss).

  \item \textbf{Backpropagation.} The error is propagated backwards
    through the entire network. Crucially, this includes the
    embedding table: the gradient tells the model \emph{how to adjust
    the embedding vector} of each token that appeared in the input to
    reduce the error.

  \item \textbf{Repeat billions of times.} Over billions of training
    examples, tokens that appear in similar contexts (``The \rule{1cm}{0.15mm}
    sat on the mat,'' ``I fed the \rule{1cm}{0.15mm}'') get their
    embedding vectors pushed in similar directions. Gradually, ``cat''
    and ``dog'' drift towards each other in the embedding space, while
    ``cat'' and ``bicycle'' drift apart.
\end{enumerate}

The resulting structure is called \concept{distributional
semantics}---the meaning of a word is defined by the company it
keeps. Nobody told the model that ``cat'' and ``dog'' are both
animals; it discovered this from their overlapping usage patterns.

\begin{engineernote}
The embedding table is a \emph{learned parameter} of the model, just
like the weight matrices $\bm{W}_Q$, $\bm{W}_K$, $\bm{W}_V$ in
self-attention or $\bm{W}_1$, $\bm{W}_2$ in the MLP. The difference
is that the embedding lookup is not a matrix multiplication---it is a
simple table read (select row $i$). This makes it the one operation
in the entire model that is already ``bitwise-friendly'': no
floating-point arithmetic, just a memory access.
\end{engineernote}

\subsubsection{Why High Dimensions Matter}

Two dimensions can separate cats from cars. But to simultaneously
encode that ``bank'' can mean a financial institution \emph{or} a
river bank, that ``run'' can be a verb \emph{or} a noun, and that
``Java'' can be a programming language \emph{or} an island---you need
many dimensions. Each dimension can encode a different aspect of
meaning, and a token's position in this space encodes all its
potential meanings simultaneously.

This is the first point of tension with bitwise approaches:
\emph{embeddings are inherently continuous}. The difference between
``happy'' and ``joyful'' is a small vector in this space; booleanising
it (rounding each dimension to 0 or~1) would destroy these fine
distinctions.

\subsubsection{But What About Subword Pieces?}

If embeddings represent meaning, and the vocabulary consists of
subword fragments, a natural question arises: how can a word like
``prince''---split by BPE into ``prin'' + ``ce''---end up with a
meaningful representation? Neither ``prin'' nor ``ce'' means
``prince'' on its own.

The answer is that the embedding table only provides the
\emph{raw materials}. The transformer layers that follow are what
\emph{compose} meaning from fragments:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Embedding lookup.} The tokens ``prin'' and ``ce''
    each get their own vector from the embedding table. At the start
    of training, these are random; after training, they carry partial
    information (``prin'' appears in ``prince,'' ``print,''
    ``principal''---its embedding encodes what these contexts share).

  \item \textbf{Self-attention assembles the word.} In the first
    transformer layer, ``prin'' attends to ``ce'' and vice versa.
    The model has learned from billions of examples that these two
    tokens together form a concept. Their representations are updated
    to incorporate each other's information.

  \item \textbf{Deeper layers build full meaning.} After a few
    layers, the vector that started as ``prin'' effectively represents
    ``prince''---it has absorbed the context from ``ce'' through
    attention. By the deeper layers, the representations are no longer
    about subword pieces at all; they represent the \emph{contextual
    meaning} of the full word in its sentence.
\end{enumerate}

\begin{keyinsight}
The model does not distinguish between ``assembling a word from
subword pieces'' and ``assembling a sentence from words.'' Both are
solved by the same self-attention mechanism. This is why LLMs handle
misspellings, novel words, and even code so gracefully: ``princ'' +
``ess'' works the same way as ``prin'' + ``ce.'' The pieces carry
partial information, and attention composes them into meaning---layer
by layer, from fragments to words to phrases to full understanding.
\end{keyinsight}

% ----------------------------------------------------------------------------
\subsection{The Transformer Block: The Engine of Modern AI}
\label{sec:transformer-block}
% ----------------------------------------------------------------------------

The transformer~\cite{vaswani2017attention} is the architecture
behind essentially all modern LLMs (GPT, Claude, Llama, etc.). A
transformer is a stack of identical \concept{transformer blocks}, each
containing two main sub-layers:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Self-attention}: lets each token look at every other
    token and decide which ones are relevant.
  \item \textbf{Feed-forward network (MLP)}: processes each token
    independently through a small neural network.
\end{enumerate}

Each sub-layer is wrapped in a \concept{residual connection} (add the
input back to the output) and \concept{layer normalisation} (rescale
the values to prevent them from growing too large or too small).

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  block/.style={draw, rounded corners=3pt, minimum width=4cm,
    minimum height=0.8cm, font=\small\sffamily, align=center},
  arr/.style={-{Stealth[length=5pt]}, thick},
  plus/.style={circle, draw, minimum size=0.5cm, font=\small,
    fill=bitgreen!10, inner sep=0pt},
  skip/.style={thick, bitgray, dashed}
]
  % Input
  \node[font=\small\sffamily, bitgray] (in) at (0, 0) {token embeddings};

  % Self-attention
  \node[block, fill=bitblue!10, draw=bitblue] (attn)
    at (0, 1.5) {Self-Attention};
  \node[plus] (add1) at (0, 2.8) {$+$};
  \node[block, fill=bitgray!10, draw=bitgray] (ln1)
    at (0, 3.8) {Layer Normalisation};

  % MLP
  \node[block, fill=bitorange!10, draw=bitorange] (mlp)
    at (0, 5.0) {Feed-Forward Network (MLP)};
  \node[plus] (add2) at (0, 6.3) {$+$};
  \node[block, fill=bitgray!10, draw=bitgray] (ln2)
    at (0, 7.3) {Layer Normalisation};

  % Output
  \node[font=\small\sffamily, bitgray] (out) at (0, 8.5)
    {to next block};

  % Main flow
  \draw[arr] (in) -- (attn);
  \draw[arr] (attn) -- (add1);
  \draw[arr] (add1) -- (ln1);
  \draw[arr] (ln1) -- (mlp);
  \draw[arr] (mlp) -- (add2);
  \draw[arr] (add2) -- (ln2);
  \draw[arr] (ln2) -- (out);

  % Residual connections
  \draw[skip] (-2.8, 0.3) -- (-2.8, 2.8);
  \draw[skip, -{Stealth[length=4pt]}] (-2.8, 2.8) -- (add1);
  \draw[skip] (-2.8, 4.1) -- (-2.8, 6.3);
  \draw[skip, -{Stealth[length=4pt]}] (-2.8, 6.3) -- (add2);

  % Labels
  \node[font=\scriptsize\sffamily, bitgray, left] at (-3.2, 1.5)
    {residual};
  \node[font=\scriptsize\sffamily, bitgray, left] at (-3.2, 5.2)
    {residual};

  % Bracket for "one block"
  \draw[decorate, decoration={brace, amplitude=8pt, mirror},
    thick, bitpurple]
    (3, 0.7) -- (3, 8.0);
  \node[font=\small\sffamily, bitpurple, right, text width=3cm]
    at (3.5, 4.3)
    {one transformer\\block\\[4pt]\scriptsize (repeated\\32--80 times)};
\end{tikzpicture}
\caption{A single transformer block. The self-attention sub-layer lets
  tokens communicate with each other; the MLP sub-layer processes each
  token independently. Residual connections and layer normalisation
  stabilise training. A full LLM stacks 32--80 of these blocks.}
\label{fig:transformer-block}
\end{figure}

\begin{analogy}
Think of a transformer block as a two-step meeting process. In the
\emph{self-attention} step, every participant in the room looks
around, identifies who has relevant information, and takes notes.
In the \emph{MLP} step, each participant goes back to their desk
and privately thinks about what they learned. Then the whole process
repeats---another round of group discussion followed by private
reflection. After 32--80 rounds, each participant has a rich
understanding of the entire conversation.

The residual connection is like keeping your original notes alongside
the new ones---you never lose what you started with. Layer
normalisation is like agreeing on a common scale for everyone's
notes so that one loud participant does not dominate.
\end{analogy}

Let us examine each component in detail.

% ----------------------------------------------------------------------------
\subsection{Self-Attention: The Core Innovation}
\label{sec:self-attention}
% ----------------------------------------------------------------------------

Self-attention is the mechanism that lets each token in a sequence
\emph{attend to} every other token. But what does ``attend to''
actually mean in practice? It means: \textbf{compute a relevance score
against every other token, then blend their information proportionally.}
The result is a new representation of the token that now \emph{contains}
information from the tokens it found relevant.

\subsubsection{The Intuition}

Consider the sentence: ``The animal didn't cross the street because
it was too wide.'' What does ``it'' refer to? A human instantly knows
``it'' means ``the street'' (streets are wide, not animals). But how
does a neural network figure this out?

Self-attention lets the token ``it'' \emph{look at every other token}
and assign each one a relevance score. It assigns high relevance to
``street'' and low relevance to ``animal,'' then uses these scores to
blend information from the relevant tokens into its own representation.
Concretely, this is what happens for a single token:

\begin{enumerate}[leftmargin=2em]
  \item The token produces a \emph{query} vector---a question:
    ``what kind of context do I need?''
  \item Every token (including itself) produces a \emph{key}
    vector---a label: ``here is what I contain.''
  \item The query is dot-producted with every key, giving a raw
    similarity score per token. High dot product = ``this token is
    relevant to me.''
  \item The scores are normalised via softmax so they sum to~1
    (now they are weights, e.g.\ 0.02, 0.71, 0.05, \ldots).
  \item Every token also produces a \emph{value} vector---the actual
    information it offers if selected.
  \item The value vectors are combined as a weighted sum using the
    attention weights. So ``it'' gets 71\% of ``cat'''s
    information, 19\% of its own, 2\% of ``The,'' and so on.
\end{enumerate}

This happens \emph{for every token simultaneously}---every token
queries every other token in parallel, producing an $n \times n$
attention matrix. That is why it is called \emph{self}-attention (the
sequence attends to itself) and why it costs $O(n^2)$ in sequence
length.

\begin{keyinsight}
The number of tokens that self-attention can look at is exactly what
defines a model's \concept{context window}. In standard self-attention,
every token attends to \emph{every other token} in the
sequence---there is no built-in limit on how far back it looks. The
only constraint is how many tokens the model was trained to handle
at once.

This is why the context window has become one of the most important
model parameters: a longer context means self-attention can draw on
more information, but the $O(n^2)$ cost grows rapidly.
Early transformers were limited to 512 or 2{,}048 tokens.
Modern LLMs have pushed this dramatically:
Llama~3 supports up to 128K tokens, GPT-4 handles 128K, and
Claude operates with up to 1{,}000{,}000 tokens---requiring
significant engineering to manage the quadratic attention cost at
that scale (techniques like sparse attention, sliding windows, and
efficient memory management).

Some architectures deliberately limit the look-back to a fixed
window (e.g.\ Mistral uses a 4{,}096-token sliding window per
layer) to reduce cost, relying on information propagating across layers to
effectively extend the range. But the standard mechanism attends to
the full context.
\end{keyinsight}

\subsubsection{The Mechanism: Queries, Keys, and Values}

Self-attention works through three learned projections. Each token's
embedding vector $\bm{x} \in \mathbb{R}^d$ is multiplied by three
different weight matrices to produce three different vectors:

\begin{itemize}[leftmargin=2em]
  \item A \concept{query} ($\bm{q} = \bm{x}\bm{W}_Q$): ``What am I looking for?''
  \item A \concept{key} ($\bm{k} = \bm{x}\bm{W}_K$): ``What do I contain?''
  \item A \concept{value} ($\bm{v} = \bm{x}\bm{W}_V$): ``What information do I
    provide if selected?''
\end{itemize}

All three come from the \emph{same} embedding vector---but the learned
weight matrices $\bm{W}_Q$, $\bm{W}_K$, $\bm{W}_V$ project it into
different subspaces, each serving a different role. The key--value
split is essential: the key determines \emph{whether} a token is
relevant (used for scoring), while the value determines \emph{what
information} gets passed along (used for blending). They are
deliberately decoupled so the network can learn to match on one
feature (e.g.\ grammatical role) but transmit another (e.g.\ semantic
content).

\begin{engineernote}
Why does the dot product measure similarity? Two vectors pointing in
the same direction have a large positive dot product; perpendicular
vectors have a dot product of zero; opposing vectors are negative.
So $\bm{q} \cdot \bm{k} = \sum_i q_i k_i$ is large when query and
key ``agree'' across many dimensions---exactly the property we need
for a relevance score. Training adjusts the projection matrices so
that tokens that should attend to each other end up with aligned
query and key vectors.
\end{engineernote}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  tok/.style={draw, rounded corners=2pt, minimum width=1.2cm,
    minimum height=0.6cm, font=\scriptsize\sffamily, fill=#1!10},
  mat/.style={draw, rounded corners=2pt, minimum width=1cm,
    minimum height=0.6cm, font=\scriptsize\sffamily, fill=#1!15},
  arr/.style={-{Stealth[length=4pt]}, thick},
  score/.style={font=\scriptsize\ttfamily}
]
  % Tokens
  \foreach \i/\w in {1/The, 2/cat, 3/sat, 4/on, 5/it} {
    \node[tok=bitgray] (t\i) at ({(\i-1)*2.5}, 0) {\w};
  }

  % Query from "it"
  \node[mat=bitblue, minimum width=1.5cm] (q) at (10, 1.5)
    {query: ``it''};
  \draw[arr, bitblue] (t5.north) -- (q.south);

  % Keys from all
  \foreach \i/\w in {1/The, 2/cat, 3/sat, 4/on, 5/it} {
    \node[mat=bitgreen] (k\i) at ({(\i-1)*2.5}, 2.5)
      {key: ``\w''};
    \draw[arr, bitgreen] (t\i.north) -- (k\i.south);
  }

  % Attention scores
  \node[font=\small\sffamily, bitgray] at (5, 4) {attention scores
    (query $\cdot$ key):};
  \foreach \i/\s/\h in {1/0.02/2, 2/0.71/35, 3/0.05/3, 4/0.03/2, 5/0.19/10} {
    \fill[bitblue!\h] ({(\i-1)*2.5-0.4}, 4.5) rectangle
      ({(\i-1)*2.5+0.4}, {4.5+\s*1.5});
    \node[score] at ({(\i-1)*2.5}, {4.5+\s*1.5+0.3}) {\s};
  }

  % Label
  \node[font=\scriptsize\sffamily, bitgray, text width=4cm, align=left]
    at (12.5, 3.5) {``it'' attends\\most strongly\\to ``cat''\\($\leftarrow$ correct\\referent)};
\end{tikzpicture}
\caption{Self-attention for the token ``it''. Its query is compared
  (via dot product) with every token's key to produce attention scores.
  The highest score (0.71) goes to ``cat''---the token ``it'' refers to.
  These scores then weight each token's value to produce the output.}
\label{fig:attention-scores}
\end{figure}

\subsubsection{The Mathematics}

Given a sequence of $n$ token embeddings packed into a matrix
$\bm{X} \in \mathbb{R}^{n \times d}$, self-attention computes:

\begin{align}
  \bm{Q} &= \bm{X}\bm{W}_Q, \quad
  \bm{K} = \bm{X}\bm{W}_K, \quad
  \bm{V} = \bm{X}\bm{W}_V
  \label{eq:qkv}\\[6pt]
  \text{Attention}(\bm{Q}, \bm{K}, \bm{V})
  &= \underbrace{\text{softmax}\!\left(
    \frac{\bm{Q}\bm{K}^\top}{\sqrt{d_k}}
  \right)}_{\text{attention weights}}
  \bm{V}
  \label{eq:attention}
\end{align}

where $\bm{W}_Q, \bm{W}_K, \bm{W}_V \in \mathbb{R}^{d \times d_k}$
are learned projection matrices, and $d_k$ is the dimension of each
head (typically $d / h$ where $h$ is the number of attention heads).

Let us unpack each step:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Project.} Multiply the input by three weight matrices
    to get queries, keys, and values. This is standard matrix
    multiplication---the same MAC operation we have been trying to
    eliminate.

  \item \textbf{Score.} Compute $\bm{Q}\bm{K}^\top$: the dot product
    of every query with every key. This produces an $n \times n$
    matrix of raw relevance scores. For a sequence of 4{,}096 tokens,
    this is a $4{,}096 \times 4{,}096$ matrix---about 16~million
    scores.

  \item \textbf{Scale.} Divide by $\sqrt{d_k}$ to prevent the dot
    products from growing too large (which would push softmax into
    regions where its gradient vanishes).

  \item \textbf{Softmax.} Apply the softmax function to each row:
    \begin{equation}
      \text{softmax}(z_i) = \frac{e^{z_i}}{\sum_j e^{z_j}}
      \label{eq:softmax}
    \end{equation}
    This converts raw scores into a probability distribution: all
    positive, summing to~1. The token with the highest relevance
    gets the most weight.

  \item \textbf{Aggregate.} Multiply the attention weights by the
    value matrix. Each token's output is a weighted average of all
    tokens' values, where the weights reflect relevance. The result
    is a matrix $\in \mathbb{R}^{n \times d_k}$---one new vector per
    token, same shape as the input, but now enriched with contextual
    information from the tokens each one attended to.
\end{enumerate}

\begin{keyinsight}
The softmax function is the critical bottleneck for binarisation.
It requires:
\begin{itemize}[leftmargin=1em]
  \item Exponentiation ($e^{z_i}$)---a transcendental function that
    maps integers to irrational numbers.
  \item Division---to normalise the probabilities.
  \item Continuous outputs---the weights 0.71 and 0.72 must be
    distinguishable, not rounded to the same bit.
\end{itemize}
Every other step (projection, scoring, aggregation) is matrix
multiplication, which \emph{can} be approximated with integer or
ternary arithmetic (as BitNet demonstrates;
see \cref{sec:bitnet-reality}). The softmax is where
floating-point computation concentrates.
\end{keyinsight}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth]
  % --- Left: raw scores (bar chart) ---
  \node[font=\small\sffamily\bfseries] at (2, 4.5)
    {Raw scores (logits)};

  \foreach \i/\v/\lbl in {
    1/2.0/2.0, 2/1.0/1.0, 3/0.1/0.1, 4/-1.0/{-1.0}, 5/-3.5/{-3.5}} {
    \fill[bitblue!40] ({\i*0.7-0.1}, 0) rectangle ({\i*0.7+0.4}, {\v/2+2});
    \node[font=\tiny\sffamily] at ({\i*0.7+0.15}, {\v/2+2+0.3}) {\lbl};
  }
  \draw[->] (0.3, 0) -- (4.2, 0);
  \draw[->] (0.5, -0.2) -- (0.5, 4);
  % zero line
  \draw[gray, dashed, thin] (0.5, 2) -- (4.0, 2);
  \node[font=\tiny\sffamily, gray] at (0.2, 2) {0};

  % --- Arrow ---
  \draw[->, very thick, bitgray] (4.8, 2) -- (6.8, 2);
  \node[font=\scriptsize\sffamily, above=2pt] at (5.8, 2)
    {softmax};

  % --- Right: probabilities (bar chart) ---
  \node[font=\small\sffamily\bfseries] at (10, 4.5)
    {Probabilities (sum = 1)};

  % softmax([2.0, 1.0, 0.1, -1.0, -3.5]) ≈ [0.588, 0.216, 0.088, 0.029, 0.002]
  \foreach \i/\v/\lbl in {
    1/0.588/0.59, 2/0.216/0.22, 3/0.088/0.09,
    4/0.029/0.03, 5/0.002/{\ensuremath{<}0.01}} {
    \fill[bitorange!60] ({7+\i*0.7-0.1}, 0)
      rectangle ({7+\i*0.7+0.4}, {\v*6});
    \node[font=\tiny\sffamily] at ({7+\i*0.7+0.15}, {\v*6+0.3})
      {\lbl};
  }
  \draw[->] (7.3, 0) -- (11.5, 0);
  \draw[->] (7.5, -0.2) -- (7.5, 4);
  \node[font=\tiny\sffamily, gray, anchor=east] at (7.45, 0) {0};
  \node[font=\tiny\sffamily, gray, anchor=east] at (7.45, 3) {0.5};
  \draw[gray, dashed, thin] (7.5, 3) -- (11.3, 3);

  % Labels under bars
  \foreach \i/\lbl in {1/$z_1$, 2/$z_2$, 3/$z_3$, 4/$z_4$, 5/$z_5$} {
    \node[font=\tiny\sffamily] at ({\i*0.7+0.15}, -0.3) {\lbl};
    \node[font=\tiny\sffamily] at ({7+\i*0.7+0.15}, -0.3) {\lbl};
  }
\end{tikzpicture}
\caption{The softmax function in action. Left: raw scores can be any
  real number, including negative values. Right: after softmax, all
  values are positive and sum to~1. Notice how the largest input
  ($z_1 = 2.0$) dominates the output (59\%), while differences are
  amplified---the negative scores are nearly zeroed out. The
  exponentiation $e^{z_i}$ is what creates this winner-take-more
  effect.}
\label{fig:softmax}
\end{figure}

To see how softmax behaves continuously, consider a three-element
vector $[x,\; 1,\; 0]$ where we vary $x$ from $-4$ to $+4$.
\Cref{fig:softmax-curves} shows how the probability assigned to each
element shifts as $x$ changes: when $x$ is large, it dominates; when
$x$ is small, the other elements share the probability mass.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
  \begin{axis}[
    width=12cm, height=6.5cm,
    xmin=-4, xmax=4, ymin=0, ymax=1.05,
    axis lines=left,
    xlabel={$x$},
    ylabel={softmax probability},
    xlabel style={font=\small\sffamily},
    ylabel style={font=\small\sffamily},
    legend style={at={(0.98, 0.98)}, anchor=north east,
      font=\scriptsize\sffamily, draw=none, fill=white,
      fill opacity=0.85, text opacity=1},
    grid=major, grid style={gray!20},
    tick label style={font=\tiny},
    every axis plot/.append style={thick, samples=100, smooth},
    ytick={0, 0.2, 0.4, 0.6, 0.8, 1.0}
  ]
    % softmax([x, 1, 0]) for element x:
    % p(x) = e^x / (e^x + e^1 + e^0) = e^x / (e^x + e + 1)
    \addplot[bitblue, domain=-4:4]
      {exp(x) / (exp(x) + exp(1) + 1)};
    \addlegendentry{$P(z_1)$: element being varied ($z_1 = x$)}

    % p(z2=1) = e^1 / (e^x + e^1 + 1)
    \addplot[bitorange, domain=-4:4]
      {exp(1) / (exp(x) + exp(1) + 1)};
    \addlegendentry{$P(z_2)$: fixed element ($z_2 = 1$)}

    % p(z3=0) = 1 / (e^x + e^1 + 1)
    \addplot[bitgreen, domain=-4:4]
      {1 / (exp(x) + exp(1) + 1)};
    \addlegendentry{$P(z_3)$: fixed element ($z_3 = 0$)}

    % Reference line: equal probability
    \addplot[gray, dashed, thin, domain=-4:4, forget plot] {1/3};
    \node[font=\scriptsize\sffamily, gray, anchor=west]
      at (axis cs:2.5, 0.38)
      {$\frac{1}{3}$ (equal share)};
  \end{axis}
\end{tikzpicture}
\caption{Softmax output for a three-element vector $[x, 1, 0]$ as
  $x$ varies. When $x \ll 0$, the varied element gets near-zero
  probability and the fixed elements share the rest. When $x \gg 0$,
  it dominates. At $x = 1$, elements $z_1$ and $z_2$ are equal and
  $z_3$ gets less. The curves show the characteristic S-shaped
  transition---in fact, for a two-element softmax, this reduces
  exactly to the sigmoid function.}
\label{fig:softmax-curves}
\end{figure}

\subsubsection{Multi-Head Attention}

A single attention head can only learn one notion of ``relevance.''
But language has many simultaneous relationships: syntactic (subject
agrees with verb), semantic (synonyms cluster), positional (nearby
words matter for idioms), referential (pronouns link to antecedents).
One head cannot capture all of these at once.

The solution is \concept{multi-head attention}: instead of computing
one set of queries, keys, and values with the full embedding dimension
$d$, the model splits the embedding into $h$ smaller slices of
dimension $d_k = d/h$ and runs a \emph{separate} attention
computation on each slice, in parallel. Each slice is called an
\concept{attention head}.

Concretely, for an embedding of $d = 512$ and $h = 8$ heads:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Split.} Each token's $d$-dimensional embedding is
    projected into $h = 8$ separate query/key/value triples, each of
    dimension $d_k = 64$.
  \item \textbf{Attend independently.} Each head runs the full
    self-attention mechanism (score, scale, softmax, aggregate) on
    its own 64-dimensional slice. Head~1 might learn to track
    subject--verb agreement; Head~2 might track coreference
    (``it'' $\to$ ``street''); Head~3 might attend to nearby words.
  \item \textbf{Concatenate.} The $h$ output vectors (each 64-dim)
    are concatenated back into a single $d$-dimensional vector.
  \item \textbf{Project.} A final weight matrix $\bm{W}_O$ mixes
    the concatenated heads into the final output.
\end{enumerate}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  box/.style={draw, rounded corners=2pt, minimum height=0.7cm,
    font=\scriptsize\sffamily, fill=#1!12, align=center},
  arr/.style={->, thick},
  brace/.style={decorate, decoration={brace, amplitude=5pt, mirror}}
]
  % Input embedding
  \node[box=bitgray, minimum width=8cm] (input) at (4, 0)
    {Token embedding ($d = 512$)};

  % Heads
  \foreach \i/\col in {1/bitblue, 2/bitgreen, 3/bitorange, 4/bitgray} {
    \pgfmathsetmacro{\x}{(\i-1)*2.2 + 0.6}
    % Split box
    \node[box=\col, minimum width=1.6cm, text width=1.4cm]
      (head\i) at (\x, 2)
      {Head \i\\[-1pt]{\tiny $d_k\!=\!64$}};
    \draw[arr] (input.north -| head\i) -- (head\i.south);
    % Attention
    \node[box=\col, minimum width=1.6cm, text width=1.4cm]
      (attn\i) at (\x, 4)
      {Self-Attn\\[-1pt]{\tiny $Q_\i K_\i V_\i$}};
    \draw[arr] (head\i.north) -- (attn\i.south);
  }

  % Dots between head 4 and implied others
  \node[font=\sffamily] at (8.6, 2) {\ldots};
  \node[font=\sffamily] at (8.6, 4) {\ldots};

  % Concat
  \node[box=bitgray, minimum width=8cm] (concat) at (4, 5.8)
    {Concatenate ($h \times d_k = 512$)};
  \foreach \i in {1,...,4} {
    \draw[arr] (attn\i.north) -- (concat.south -| attn\i);
  }

  % Output projection
  \node[box=bitgray, minimum width=8cm] (output) at (4, 7.2)
    {Output projection $\bm{W}_O$ ($d = 512$)};
  \draw[arr] (concat.north) -- (output.south);
\end{tikzpicture}
\caption{Multi-head attention splits the embedding into $h$ independent
  heads, each running self-attention on a smaller slice. The results are
  concatenated and projected back to the original dimension. This lets
  the model attend to different types of relationships simultaneously.}
\label{fig:multi-head}
\end{figure}

\begin{analogy}
Multi-head attention is like having multiple reading comprehension
specialists working on the same text simultaneously. One specialist
focuses on grammar, another on meaning, a third on word proximity.
Each independently decides what is relevant, then they pool their
notes into a single summary.
\end{analogy}

The output of multi-head attention is:
\begin{equation}
  \text{MultiHead}(\bm{X}) =
  [\text{head}_1; \text{head}_2; \ldots; \text{head}_h]\bm{W}_O
\end{equation}
where each $\text{head}_i = \text{Attention}(\bm{X}\bm{W}_Q^i,
\bm{X}\bm{W}_K^i, \bm{X}\bm{W}_V^i)$ and $\bm{W}_O \in
\mathbb{R}^{d \times d}$ is a learned output projection.

% ----------------------------------------------------------------------------
\subsection{The Feed-Forward Network (MLP Block)}
\label{sec:ffn}
% ----------------------------------------------------------------------------

After attention lets tokens communicate, each token is processed
\emph{independently} through a small two-layer neural network.
``Two-layer'' here means exactly two matrix multiplications with a
nonlinear activation in between---it is the simplest possible neural
network that can learn nonlinear patterns:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Expand.} Multiply the token vector ($d$-dimensional)
    by weight matrix $\bm{W}_1$ to project it into a much larger
    hidden space ($d_{\text{ff}}$-dimensional, typically $4 \times d$).
    Then apply a nonlinear activation function $\sigma$ (typically GELU
    or SiLU). This expansion lets the network represent richer,
    more complex patterns than the original $d$ dimensions allow.
  \item \textbf{Compress.} Multiply by $\bm{W}_2$ to project back down
    to $d$ dimensions. The output has the same shape as the input, but
    the information has been transformed.
\end{enumerate}

In equation form:
\begin{equation}
  \text{FFN}(\bm{x}) = \sigma(\bm{x}\bm{W}_1 + \bm{b}_1)\bm{W}_2 + \bm{b}_2
  \label{eq:ffn}
\end{equation}

where $\sigma$ is the activation function,
$\bm{W}_1 \in \mathbb{R}^{d \times d_{\text{ff}}}$ and
$\bm{W}_2 \in \mathbb{R}^{d_{\text{ff}} \times d}$.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  box/.style={draw, rounded corners=2pt, minimum height=0.7cm,
    font=\scriptsize\sffamily, fill=#1!12, align=center},
  arr/.style={->, thick}
]
  % Input
  \node[box=bitblue, minimum width=2cm] (in) at (0, 0)
    {$\bm{x}$\\[-1pt]{\tiny $d = 512$}};

  % W1
  \node[box=bitorange, minimum width=3cm] (w1) at (3.5, 0)
    {$\times\;\bm{W}_1$\\[-1pt]{\tiny expand to $d_{\text{ff}} = 2048$}};
  \draw[arr] (in) -- (w1);

  % Activation
  \node[box=bitgreen, minimum width=1.5cm] (act) at (7, 0)
    {$\sigma$ (GELU)};
  \draw[arr] (w1) -- (act);

  % W2
  \node[box=bitorange, minimum width=3cm] (w2) at (10.5, 0)
    {$\times\;\bm{W}_2$\\[-1pt]{\tiny compress to $d = 512$}};
  \draw[arr] (act) -- (w2);

  % Output
  \node[box=bitblue, minimum width=2cm] (out) at (14, 0)
    {$\bm{y}$\\[-1pt]{\tiny $d = 512$}};
  \draw[arr] (w2) -- (out);

  % Dimension annotation
  \draw[<->, gray, thick] (3.5, -1) -- (10.5, -1);
  \node[font=\scriptsize\sffamily, gray, below=2pt] at (7, -1)
    {high-dimensional hidden space ($4\times$ wider)};
\end{tikzpicture}
\caption{The feed-forward network (MLP block): abstract pipeline view.
  The input is expanded to a wider hidden dimension, passed
  through a nonlinear activation, then compressed back.}
\label{fig:ffn}
\end{figure}

What does this look like at the level of individual neurons?
Exactly the same structure as the traditional neuron from
\cref{fig:traditional-neuron}, but arranged in two fully connected
layers (\cref{fig:ffn-neurons}). Each hidden neuron receives
\emph{all} $d$ inputs, computes a weighted sum plus bias
($\sum_i w_i x_i + b$), and applies the activation function---the
same multiply-accumulate (MAC) operation from
\cref{sec:traditional-neuron}.
The second layer does the same in reverse: each output neuron
receives all $d_{\text{ff}}$ hidden activations and produces one
component of the output vector. Every arrow in the diagram
represents one learned weight---and with $d = 512$ and
$d_{\text{ff}} = 2{,}048$, the first layer alone has
$512 \times 2{,}048 = 1{,}048{,}576$ weights.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  neuron/.style={draw, circle, minimum size=0.45cm,
    inner sep=0pt, font=\tiny\sffamily, fill=#1!15},
  arr/.style={-, thin, opacity=0.25},
  lbl/.style={font=\scriptsize\sffamily}
]
  % --- Input layer (d = 512, show 5) ---
  \foreach \i in {1,...,5} {
    \node[neuron=bitblue] (in\i) at (0, {-\i*0.8 + 0.4}) {};
  }
  \node[lbl, rotate=90, anchor=south] at (-0.7, -2)
    {Input ($d = 512$)};
  % dots
  \node[font=\small] at (0, -4.6) {$\vdots$};
  \node[neuron=bitblue] (in6) at (0, -5.2) {};

  % --- Hidden layer (d_ff = 2048, show 8) ---
  \foreach \i in {1,...,8} {
    \node[neuron=bitorange] (h\i) at (4.5, {-\i*0.65 + 0.65}) {};
  }
  \node[lbl, rotate=90, anchor=south] at (3.6, -2.5)
    {Hidden ($d_{\text{ff}} = 2048$)};
  \node[font=\small] at (4.5, -5.3) {$\vdots$};
  \node[neuron=bitorange] (h9) at (4.5, -5.9) {};

  % --- Output layer (d = 512, show 5) ---
  \foreach \i in {1,...,5} {
    \node[neuron=bitblue] (out\i) at (9, {-\i*0.8 + 0.4}) {};
  }
  \node[lbl, rotate=90, anchor=south] at (9.7, -2)
    {Output ($d = 512$)};
  \node[font=\small] at (9, -4.6) {$\vdots$};
  \node[neuron=bitblue] (out6) at (9, -5.2) {};

  % --- Connections: input -> hidden (sample) ---
  \foreach \i in {1,...,5,6} {
    \foreach \j in {1,...,8,9} {
      \draw[arr, bitorange] (in\i) -- (h\j);
    }
  }

  % --- Connections: hidden -> output (sample) ---
  \foreach \i in {1,...,8,9} {
    \foreach \j in {1,...,5,6} {
      \draw[arr, bitblue] (h\i) -- (out\j);
    }
  }

  % --- Labels ---
  \node[lbl, bitorange, align=center] at (2.25, 1.2)
    {$\bm{W}_1$: each arrow\\is one weight};
  \node[lbl, bitblue, align=center] at (6.75, 1.2)
    {$\bm{W}_2$: each arrow\\is one weight};

  % Activation annotation
  \draw[thick, bitgreen, decorate,
    decoration={brace, amplitude=4pt, mirror}]
    (5.2, 0.4) -- (5.2, -6.2);
  \node[lbl, bitgreen, anchor=west, text width=2.5cm] at (5.4, -2.9)
    {each neuron\\applies $\sigma$\\(GELU)};

  % Single neuron callout
  \draw[thick, dashed, bitgray] (3.8, -0.5) rectangle (5.2, 0.2);
  \node[lbl, bitgray, anchor=west] at (5.3, -0.15)
    {one neuron = Fig.~5};
\end{tikzpicture}
\caption{The same MLP block as \cref{fig:ffn}, drawn at the neuron
  level (cf.~\cref{fig:traditional-neuron}). Every circle is one
  neuron performing a weighted sum + activation. Every line is one
  learned weight. The input layer connects fully to a wider hidden
  layer ($\bm{W}_1$), each hidden neuron applies the activation
  function, then the hidden layer connects fully back to the output
  ($\bm{W}_2$). Only a fraction of the neurons and connections are
  shown; the real network has over one million weights in $\bm{W}_1$
  alone.}
\label{fig:ffn-neurons}
\end{figure}

The activation function $\sigma$ introduces the nonlinearity that
makes the network more than just a matrix multiplication. Without it,
stacking two linear layers ($\bm{W}_1$ then $\bm{W}_2$) would
collapse into a single linear transformation---the hidden layer would
be pointless. Modern transformers use GELU or SiLU rather than the
older ReLU, because their smooth curves give better gradient flow
during training.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
  \begin{axis}[
    width=12cm, height=6cm,
    xmin=-4, xmax=4, ymin=-1.5, ymax=4,
    axis lines=middle,
    xlabel={$x$}, ylabel={$\sigma(x)$},
    xlabel style={at=(current axis.right of origin), anchor=west,
      font=\small\sffamily},
    ylabel style={at=(current axis.above origin), anchor=south,
      font=\small\sffamily},
    legend style={at={(0.02, 0.98)}, anchor=north west,
      font=\scriptsize\sffamily, draw=none, fill=white, fill opacity=0.8,
      text opacity=1},
    grid=major, grid style={gray!20},
    tick label style={font=\tiny},
    every axis plot/.append style={thick, samples=100, smooth}
  ]
    % ReLU (two segments to get the kink)
    \addplot[bitgray, dashed, domain=-4:0, forget plot] {0};
    \addplot[bitgray, dashed, domain=0:4] {x};
    \addlegendentry{ReLU: $\max(0, x)$}

    % GELU (sigmoid approximation: x * sigmoid(1.702x))
    \addplot[bitblue, domain=-4:4]
      {x / (1 + exp(-1.702*x))};
    \addlegendentry{GELU: $x \cdot \Phi(x)$}

    % SiLU / Swish
    \addplot[bitorange, domain=-4:4]
      {x / (1 + exp(-x))};
    \addlegendentry{SiLU: $x \cdot \text{sigmoid}(x)$}
  \end{axis}
\end{tikzpicture}
\caption{Activation functions used in modern transformers. ReLU
  (dashed, for reference) has a hard kink at zero. GELU and SiLU are
  smooth approximations that allow small negative values through,
  giving better gradient flow. Both are close to linear for positive
  inputs and smoothly suppress negative inputs. The key point for
  binarisation: all three require floating-point arithmetic---they
  cannot be computed with pure bitwise operations.}
\label{fig:activations}
\end{figure}

The mathematical definitions are:

\begin{align}
  \text{ReLU}(x) &= \max(0,\, x)
  \label{eq:relu}\\[6pt]
  \text{sigmoid}(x) &= \frac{1}{1 + e^{-x}}
  \label{eq:sigmoid}\\[6pt]
  \text{SiLU}(x) &= x \cdot \text{sigmoid}(x)
    = \frac{x}{1 + e^{-x}}
  \label{eq:silu}\\[6pt]
  \text{GELU}(x) &= x \cdot \Phi(x)
  \label{eq:gelu}
\end{align}

where $\Phi(x)$ is the cumulative distribution function of the
standard normal distribution---i.e.\ the probability that a random
variable drawn from a Gaussian with mean~0 and variance~1 falls
below $x$:
\begin{equation}
  \Phi(x) = \frac{1}{2}\left[1 + \text{erf}\!\left(\frac{x}{\sqrt{2}}\right)\right]
  = \int_{-\infty}^{x} \frac{1}{\sqrt{2\pi}}\, e^{-t^2/2}\, dt
  \label{eq:phi}
\end{equation}

In practice, $\Phi(x)$ is approximated as
$\text{sigmoid}(1.702\, x)$, which avoids evaluating the error
function. This makes GELU and SiLU nearly identical in form---both
multiply $x$ by a sigmoid-like gate that smoothly transitions from~0
(for large negative $x$) to~1 (for large positive $x$). The
difference is only in the sigmoid's steepness: SiLU uses
$\text{sigmoid}(x)$ while GELU uses the slightly steeper
$\text{sigmoid}(1.702\, x)$.

The inner dimension $d_{\text{ff}}$ is typically $4 \times d$. For a
model with $d = 4{,}096$, this means $\bm{W}_1$ has $4{,}096 \times
16{,}384 = 67$~million parameters---\emph{per layer}.

\begin{engineernote}
The MLP block is where most of the model's parameters and computation
live. In a typical transformer, the MLP accounts for about two-thirds
of the total parameters and floating-point operations. This makes it
the highest-value target for binarisation: if you can make the MLP
bitwise, you have eliminated the majority of the floating-point work.
\end{engineernote}

\begin{analogy}
If self-attention is the ``group discussion'' where tokens share
information, the MLP is the ``private thinking'' where each token
processes what it learned. The attention layer moves information
\emph{between} tokens; the MLP transforms information \emph{within}
each token.
\end{analogy}

% ----------------------------------------------------------------------------
\subsection{Residual Connections and Layer Normalisation}
\label{sec:residual-layernorm}
% ----------------------------------------------------------------------------

Two seemingly minor components turn out to be essential for making
deep transformers trainable.

\subsubsection{Residual Connections}

In transformer terminology, a \emph{sub-layer} is one of the two
components inside a single transformer block: either the self-attention
or the MLP (see \cref{fig:transformer-block}). A residual connection
wraps \emph{each} sub-layer individually---the input skips past that
one sub-layer and is added back to its output:
\begin{equation}
  \bm{y} = \bm{x} + \text{SubLayer}(\bm{x})
  \label{eq:residual}
\end{equation}

So in each transformer block there are \emph{two} residual
connections: one around self-attention, one around the MLP. The skip
does not bypass the entire block or multiple layers---just one
sub-layer at a time. This means each sub-layer only needs to learn a
small \emph{correction} to its input, not the entire transformation.

Why does this matter for deep networks? A model like Llama~3 stacks
96 of these blocks, giving 192 sub-layers in total. Without residual
connections, the gradient signal would have to survive 192
back-to-back matrix multiplications during backpropagation---it would
shrink to effectively zero (the \emph{vanishing gradient} problem).
The skip connections provide a direct path for gradients to flow from
the output all the way back to the input, keeping training stable.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  box/.style={draw, rounded corners=2pt, minimum height=0.8cm,
    minimum width=2.8cm, font=\scriptsize\sffamily, fill=#1!12,
    align=center},
  arr/.style={->, thick},
  lbl/.style={font=\scriptsize\sffamily, bitgray}
]
  % --- Without residual (left) ---
  \node[font=\small\sffamily\bfseries] at (1.4, 5.5) {Without residual};

  \node[box=bitblue] (in1) at (1.4, 0) {$\bm{x}$};
  \node[box=bitorange] (sub1) at (1.4, 2) {SubLayer};
  \node[box=bitblue] (out1) at (1.4, 4) {$\bm{y} = f(\bm{x})$};

  \draw[arr] (in1) -- (sub1);
  \draw[arr] (sub1) -- (out1);

  \node[lbl, text width=2.5cm, align=center] at (1.4, -1)
    {must learn the\\entire output};

  % --- With residual (right) ---
  \node[font=\small\sffamily\bfseries] at (7.5, 5.5) {With residual};

  \node[box=bitblue] (in2) at (7.5, 0) {$\bm{x}$};
  \node[box=bitorange] (sub2) at (7.5, 2) {SubLayer};
  \node[draw, circle, inner sep=1pt, font=\scriptsize\sffamily,
    fill=bitgreen!15] (plus) at (7.5, 3.5) {$+$};
  \node[box=bitblue] (out2) at (7.5, 4.8) {$\bm{y} = \bm{x} + f(\bm{x})$};

  \draw[arr] (in2) -- (sub2);
  \draw[arr] (sub2) -- (plus);
  \draw[arr] (plus) -- (out2);

  % Skip connection
  \draw[arr, bitgreen, thick] (in2.east) -- ++(1.5, 0) |- (plus.east);
  \node[lbl, bitgreen, rotate=90, anchor=south] at (9.3, 1.75)
    {skip connection};

  \node[lbl, text width=2.5cm, align=center] at (7.5, -1)
    {only learns the\\small \emph{difference}};
\end{tikzpicture}
\caption{Residual connections. Without the skip connection (left), the
  sub-layer must learn the complete output. With the skip connection
  (right), the input passes straight through and the sub-layer only
  needs to learn a small correction. This keeps gradients healthy
  even through 96+ stacked layers.}
\label{fig:residual}
\end{figure}

\subsubsection{Layer Normalisation}

Layer normalisation~\cite{ba2016layernorm} rescales each token's
representation to have zero mean and unit variance:
\begin{equation}
  \text{LayerNorm}(\bm{x}) =
  \frac{\bm{x} - \mu}{\sqrt{\sigma^2 + \epsilon}} \odot \bm{\gamma} + \bm{\beta}
  \label{eq:layernorm}
\end{equation}
where $\mu$ and $\sigma^2$ are the mean and variance computed across
the embedding dimensions of a single token, $\bm{\gamma}$ and
$\bm{\beta}$ are learned scale and shift parameters, and $\epsilon$
is a small constant for numerical stability.

\begin{engineernote}
Layer normalisation is another floating-point bottleneck. It requires:
\begin{itemize}[leftmargin=1em]
  \item Computing mean and variance (requires division and square root).
  \item The division $(\bm{x} - \mu) / \sqrt{\sigma^2 + \epsilon}$
    produces continuous values even if $\bm{x}$ is integer.
  \item The learned parameters $\bm{\gamma}$ and $\bm{\beta}$ are
    continuous.
\end{itemize}
However, approximations exist: \concept{RMSNorm} (used in Llama and
many modern LLMs) drops the mean subtraction and simplifies to
$\bm{x} / \text{RMS}(\bm{x}) \cdot \bm{\gamma}$, which is slightly
simpler. Integer approximations of RMSNorm have been demonstrated
in quantisation research, though with some accuracy trade-off
(\cref{sec:layernorm-problem}).
\end{engineernote}

% ----------------------------------------------------------------------------
\subsection{Autoregressive Generation: One Token at a Time}
\label{sec:autoregressive}
% ----------------------------------------------------------------------------

An LLM generates text through \concept{autoregressive generation}:
it predicts the next token, appends it to the sequence, and repeats.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  tok/.style={draw, rounded corners=2pt, minimum height=0.6cm,
    font=\scriptsize\sffamily, inner sep=3pt},
  pred/.style={tok, fill=bitgreen!20, draw=bitgreen},
  ctx/.style={tok, fill=bitblue!10, draw=bitblue},
  arr/.style={-{Stealth[length=4pt]}, thick, bitgray}
]
  % Step 1
  \node[font=\scriptsize\sffamily\bfseries, left] at (-0.5, 3)
    {Step 1:};
  \node[ctx] (s1a) at (1.5, 3) {The};
  \node[ctx] (s1b) at (3, 3) {cat};
  \node[pred] (s1c) at (4.5, 3) {sat};
  \draw[arr] (s1b) -- (s1c);

  % Step 2
  \node[font=\scriptsize\sffamily\bfseries, left] at (-0.5, 2)
    {Step 2:};
  \node[ctx] (s2a) at (1.5, 2) {The};
  \node[ctx] (s2b) at (3, 2) {cat};
  \node[ctx] (s2c) at (4.5, 2) {sat};
  \node[pred] (s2d) at (6, 2) {on};
  \draw[arr] (s2c) -- (s2d);

  % Step 3
  \node[font=\scriptsize\sffamily\bfseries, left] at (-0.5, 1)
    {Step 3:};
  \node[ctx] (s3a) at (1.5, 1) {The};
  \node[ctx] (s3b) at (3, 1) {cat};
  \node[ctx] (s3c) at (4.5, 1) {sat};
  \node[ctx] (s3d) at (6, 1) {on};
  \node[pred] (s3e) at (7.5, 1) {the};
  \draw[arr] (s3d) -- (s3e);

  % Step 4
  \node[font=\scriptsize\sffamily\bfseries, left] at (-0.5, 0)
    {Step 4:};
  \node[ctx] (s4a) at (1.5, 0) {The};
  \node[ctx] (s4b) at (3, 0) {cat};
  \node[ctx] (s4c) at (4.5, 0) {sat};
  \node[ctx] (s4d) at (6, 0) {on};
  \node[ctx] (s4e) at (7.5, 0) {the};
  \node[pred] (s4f) at (9, 0) {mat};
  \draw[arr] (s4e) -- (s4f);

  % Legend
  \node[ctx, right] at (10.5, 3) {context};
  \node[pred, right] at (10.5, 2.2) {predicted};
\end{tikzpicture}
\caption{Autoregressive generation. At each step, the model sees all
  previous tokens (the context) and predicts the next one. The
  predicted token is appended to the context, and the process repeats.
  Generating a 1{,}000-token response requires 1{,}000 forward passes.}
\label{fig:autoregressive}
\end{figure}

At each step, the model processes the entire context through all
transformer blocks and produces a probability distribution over the
vocabulary:

\begin{equation}
  P(t_{n+1} \mid t_1, \ldots, t_n) =
  \text{softmax}(\bm{h}_n \bm{W}_{\text{vocab}})
  \label{eq:next-token}
\end{equation}

where $\bm{h}_n$ is the hidden state of the last token after the
final transformer block, and $\bm{W}_{\text{vocab}} \in
\mathbb{R}^{d \times V}$ projects it into a score for each vocabulary
item. Let us unpack what this means step by step, because this is
where all the abstract vectors finally become concrete text.

\subsubsection{From Vector to Word: The Output Head}

After passing through all transformer blocks, the model has a single
$d$-dimensional vector $\bm{h}_n$ for the last token in the
sequence. This vector encodes everything the model ``knows'' about
what should come next. But it is still just a list of, say, 4{,}096
floating-point numbers---not a word. How do we get from here to an
actual token?

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Final layer norm.} The vector is normalised one last
    time to stabilise its scale.

  \item \textbf{Vocabulary projection.} The normalised vector
    $\bm{h}_n \in \mathbb{R}^d$ is multiplied by the
    \concept{output weight matrix} $\bm{W}_{\text{vocab}} \in
    \mathbb{R}^{d \times V}$, where $V$ is the vocabulary size
    (e.g.\ 32{,}000 for Llama, 100{,}000+ for GPT-4). This single
    matrix multiplication produces one raw score (called a
    \emph{logit}) for every token in the vocabulary:
    \[
      \text{logits} = \bm{h}_n \, \bm{W}_{\text{vocab}}
      \in \mathbb{R}^V
    \]
    Each column of $\bm{W}_{\text{vocab}}$ is essentially a
    $d$-dimensional ``template'' for one vocabulary token. The dot
    product $\bm{h}_n \cdot \bm{w}_{\text{token}}$ measures how
    well the model's hidden state matches that token's
    template---exactly the same dot-product-as-similarity idea from
    self-attention (\cref{sec:self-attention}).

  \item \textbf{Softmax.} The $V$ logits are passed through
    softmax to produce a probability distribution. For a vocabulary
    of 32{,}000 tokens, this gives 32{,}000 probabilities that sum
    to~1. The token ``cat'' might get probability 0.12, ``dog'' might
    get 0.08, and most tokens get near-zero probability.

  \item \textbf{Sample or pick.} A token is selected from this
    distribution. The simplest strategy is \emph{greedy decoding}:
    always pick the highest-probability token. In practice, models
    use \emph{sampling with temperature} to introduce controlled
    randomness, which produces more natural and varied text.
\end{enumerate}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth,
  box/.style={draw, rounded corners=2pt, minimum height=0.7cm,
    font=\scriptsize\sffamily, fill=#1!12, align=center},
  arr/.style={->, thick}
]
  % Hidden state
  \node[box=bitblue, minimum width=2.5cm] (h) at (0, 0)
    {$\bm{h}_n$\\[-1pt]{\tiny $d = 4096$}};

  % W_vocab multiply
  \node[box=bitorange, minimum width=3.5cm] (w) at (4.5, 0)
    {$\times\;\bm{W}_{\text{vocab}}$\\[-1pt]
     {\tiny $d \times V = 4096 \times 32000$}};
  \draw[arr] (h) -- (w);

  % Logits
  \node[box=bitgray, minimum width=2.5cm] (logits) at (9, 0)
    {logits\\[-1pt]{\tiny $V = 32000$ scores}};
  \draw[arr] (w) -- (logits);

  % Softmax
  \node[box=bitgreen, minimum width=2cm] (soft) at (12.5, 0)
    {softmax};
  \draw[arr] (logits) -- (soft);

  % Probabilities below
  \node[box=bitorange, minimum width=2.5cm] (probs) at (12.5, -1.8)
    {probabilities\\[-1pt]{\tiny 32000 values, sum = 1}};
  \draw[arr] (soft) -- (probs);

  % Example tokens
  \node[font=\tiny\sffamily, anchor=west, text width=4cm] at (9, -1.8)
    {``cat'' $\to$ 0.12\\
     ``dog'' $\to$ 0.08\\
     ``the'' $\to$ 0.05\\
     ``sat'' $\to$ 0.04\\
     \ldots\\[-2pt]
     32{,}000 entries};

  % Selected token
  \node[box=bitblue, minimum width=2cm] (token) at (12.5, -3.5)
    {``cat''\\[-1pt]{\tiny selected token}};
  \draw[arr] (probs) -- (token);
  \node[font=\tiny\sffamily, bitgray, anchor=west] at (13.7, -2.65)
    {sample};
\end{tikzpicture}
\caption{The output head: from hidden state to predicted token. The
  $d$-dimensional vector is projected to $V$ vocabulary scores (one
  per possible token), softmax converts these to probabilities, and a
  token is selected. The matrix $\bm{W}_{\text{vocab}}$ alone has
  $d \times V \approx 131$~million parameters.}
\label{fig:output-head}
\end{figure}

\begin{engineernote}
The vocabulary is not a dictionary of English words. It is a set of
\emph{subword tokens} learned from data by algorithms like BPE
(Byte-Pair Encoding). Common words like ``the'' are single tokens;
rarer words are split into pieces (``unforgettable'' $\to$ ``un'' +
``forget'' + ``table''). This is why LLMs can handle any text,
including code, URLs, and languages they were not primarily trained
on---the vocabulary covers all possible byte sequences, just at
different granularities.
\end{engineernote}

\begin{keyinsight}
Autoregressive generation means the model's speed is determined by
\emph{latency per token}, not throughput. Even if you can process a
batch of tokens quickly (the ``prefill'' phase), generation is
inherently sequential: you cannot predict token~$n{+}1$ until you
have produced token~$n$. This is why LLM inference is
\emph{memory-bandwidth-bound}, not compute-bound---the bottleneck is
reading the model's weights from memory for each generated token, not
the arithmetic itself. This memory-bandwidth bottleneck is exactly
where bitwise models with smaller weights have the most to offer.
\end{keyinsight}

\subsubsection{The Causal Mask}

During generation, a token must only attend to tokens \emph{before}
it---it cannot look ahead at tokens that have not been generated yet.
This is enforced by a \concept{causal mask}: the attention scores for
future positions are set to $-\infty$ before the softmax, which drives
their attention weights to zero.

\begin{equation}
  \text{CausalAttention}(\bm{Q}, \bm{K}, \bm{V}) =
  \text{softmax}\!\left(
    \frac{\bm{Q}\bm{K}^\top}{\sqrt{d_k}} + \bm{M}
  \right)\bm{V}
\end{equation}

where $M_{ij} = 0$ if $i \geq j$ and $M_{ij} = -\infty$ otherwise.

% ----------------------------------------------------------------------------
\subsection{Putting It All Together: The Full LLM}
\label{sec:full-llm}
% ----------------------------------------------------------------------------

A complete LLM is assembled from the components we have described:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  comp/.style={draw, rounded corners=3pt, minimum width=5cm,
    minimum height=0.7cm, font=\small\sffamily, align=center},
  arr/.style={-{Stealth[length=5pt]}, thick},
  brace/.style={decorate, decoration={brace, amplitude=6pt, mirror}}
]
  % Stack
  \node[comp, fill=bitgray!8] (tok) at (0, 0)
    {Tokeniser: text $\to$ [token IDs]};
  \node[comp, fill=bitpurple!8] (emb) at (0, 1.2)
    {Embedding table: [IDs] $\to$ [vectors]};
  \node[comp, fill=bitblue!8] (b1) at (0, 2.4)
    {Transformer block 1};
  \node[comp, fill=bitblue!8] (b2) at (0, 3.4)
    {Transformer block 2};
  \node[font=\sffamily, bitgray] (dots) at (0, 4.2) {$\vdots$};
  \node[comp, fill=bitblue!8] (bn) at (0, 5.0)
    {Transformer block $N$};
  \node[comp, fill=bitgray!8] (ln) at (0, 6.2)
    {Final layer norm};
  \node[comp, fill=bitgreen!8] (head) at (0, 7.4)
    {Output head: vector $\to$ vocabulary scores};
  \node[comp, fill=bitorange!8] (sm) at (0, 8.6)
    {Softmax $\to$ next token probability};

  % Arrows
  \draw[arr] (tok) -- (emb);
  \draw[arr] (emb) -- (b1);
  \draw[arr] (b1) -- (b2);
  \draw[arr] (b2) -- (dots);
  \draw[arr] (dots) -- (bn);
  \draw[arr] (bn) -- (ln);
  \draw[arr] (ln) -- (head);
  \draw[arr] (head) -- (sm);

  % Annotations
  \draw[brace, thick, bitblue]
    (3.3, 2.1) -- (3.3, 5.3);
  \node[font=\scriptsize\sffamily, bitblue, right, text width=3.5cm]
    at (3.8, 3.7)
    {$N$ = 32 (7B model)\\$N$ = 80 (70B model)\\$N$ = 126 (405B model)};

  % Operation counts
  \node[font=\scriptsize\sffamily, bitgray, left, text width=3cm,
    align=right] at (-3.5, 1.2) {lookup only\\(no FP)};
  \node[font=\scriptsize\sffamily, bitgray, left, text width=3cm,
    align=right] at (-3.5, 3.7) {matrix multiply\\+ softmax\\+ layer norm};
  \node[font=\scriptsize\sffamily, bitgray, left, text width=3cm,
    align=right] at (-3.5, 8.6) {softmax\\(continuous)};
\end{tikzpicture}
\caption{The complete LLM architecture. Token IDs enter at the bottom,
  pass through $N$ transformer blocks, and produce a probability
  distribution over the next token at the top. The vast majority of
  computation happens in the transformer blocks (middle).}
\label{fig:full-llm}
\end{figure}

\Cref{tab:llm-sizes} shows the scale of modern LLMs.

\begin{table}[htbp]
\centering
\caption{Scale of representative LLMs. Parameter counts determine
  memory requirements; token counts determine the knowledge absorbed
  during training.}
\label{tab:llm-sizes}
\begin{tabular}{@{}lrrrrr@{}}
  \toprule
  \textbf{Model} & \textbf{Params} & \textbf{Layers} & \textbf{$d$}
    & \textbf{Heads} & \textbf{Training tokens} \\
  \midrule
  GPT-2       & 1.5B    & 48  & 1{,}600  & 25  & 40B \\
  Llama 2 7B  & 7B      & 32  & 4{,}096  & 32  & 2T \\
  Llama 2 70B & 70B     & 80  & 8{,}192  & 64  & 2T \\
  Llama 3 405B& 405B    & 126 & 16{,}384 & 128 & 15T \\
  \bottomrule
\end{tabular}
\end{table}

\begin{engineernote}
A 7-billion-parameter model stored in 16-bit floating point requires
$7 \times 10^9 \times 2 = 14$\,GB of memory---just for the weights.
Add activations, key-value cache, and optimizer state, and training
requires 4--8$\times$ more. This is why bitwise models are so
appealing: ternary weights ($\{-1, 0, +1\}$) need only 1.58~bits
per parameter, reducing the 7B model from 14\,GB to under 2\,GB.
\end{engineernote}

% ----------------------------------------------------------------------------
\subsection{Mixture of Experts: Scaling Without Proportional Cost}
\label{sec:moe}
% ----------------------------------------------------------------------------

As models grow larger, a key insight emerged: not every token needs
every parameter. \concept{Mixture of Experts} (MoE) replaces the
single MLP block in each transformer layer with multiple ``expert''
MLPs, and a small \concept{gating network} (also called a router)
decides which experts process each token.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  expert/.style={draw, rounded corners=3pt, minimum width=1.8cm,
    minimum height=1.2cm, font=\scriptsize\sffamily, align=center,
    fill=#1!10, draw=#1},
  gate/.style={draw, rounded corners=3pt, minimum width=2.5cm,
    minimum height=0.8cm, font=\small\sffamily, fill=bitgray!10},
  arr/.style={-{Stealth[length=4pt]}, thick},
  sel/.style={arr, bitgreen, line width=1.5pt},
  dim/.style={arr, bitgray!40, dashed}
]
  % Input
  \node[font=\small\sffamily] (in) at (5, -0.5) {token $\bm{x}$};

  % Gate
  \node[gate] (g) at (5, 1.2) {Router / Gate};
  \draw[arr] (in) -- (g);

  % Experts
  \node[expert=bitblue]   (e1) at (0,  3.5) {Expert 1\\(MLP)};
  \node[expert=bitgreen]  (e2) at (2.5, 3.5) {Expert 2\\(MLP)};
  \node[expert=bitorange] (e3) at (5,  3.5) {Expert 3\\(MLP)};
  \node[expert=bitpurple] (e4) at (7.5, 3.5) {Expert 4\\(MLP)};
  \node[expert=bitgray]   (e5) at (10, 3.5) {Expert 5\\(MLP)};

  % Gate to experts
  \draw[dim] (g.north) -- (e1.south);
  \draw[sel] (g.north) -- (e2.south);
  \draw[dim] (g.north) -- (e3.south);
  \draw[sel] (g.north) -- (e4.south);
  \draw[dim] (g.north) -- (e5.south);

  % Labels for selected
  \node[font=\scriptsize\sffamily, bitgreen, above right] at (e2.north)
    {selected};
  \node[font=\scriptsize\sffamily, bitgreen, above right] at (e4.north)
    {selected};

  % Output
  \node[font=\small\sffamily] (out) at (5, 5.5)
    {weighted sum of expert outputs};
  \draw[arr] (e2.north) -- (out.south west);
  \draw[arr] (e4.north) -- (out.south east);

  % Annotation
  \node[font=\scriptsize\sffamily, bitgray, text width=4cm, align=left]
    at (12, 2.5) {Top-$k$ routing:\\each token uses\\only 2 of 5 experts\\($\sim$40\% of params\\activated per token)};
\end{tikzpicture}
\caption{Mixture of Experts (MoE). A gating network routes each
  token to the top-$k$ experts (typically $k=2$). The model has many
  parameters (all experts combined) but each token only activates a
  fraction of them, keeping per-token computation manageable.}
\label{fig:moe}
\end{figure}

The gating function is typically a simple linear layer followed by
softmax:
\begin{equation}
  G(\bm{x}) = \text{softmax}(\bm{x}\bm{W}_g)
  \label{eq:gating}
\end{equation}
where $\bm{W}_g \in \mathbb{R}^{d \times E}$ and $E$ is the number
of experts. The top-$k$ values of $G(\bm{x})$ determine which experts
are activated and how their outputs are weighted.

\begin{keyinsight}
MoE decouples \emph{total model knowledge} from \emph{per-token
computation}. A model can have 400B total parameters but only activate
20B per token. This is relevant for bitwise approaches because:
\begin{itemize}[leftmargin=1em]
  \item The router/gating network is a small classification problem
    (``which expert should handle this token?'')---exactly the kind
    of task where Tsetlin Machines or Logic Gate Networks could excel.
  \item Each expert MLP is independent and could be individually
    compiled to logic gates or LUTs.
  \item The ``sparse activation'' principle aligns with bitwise
    networks' strength: small, cache-resident models that run fast.
\end{itemize}
\end{keyinsight}

\begin{analogy}
MoE is like a hospital with specialist departments. A patient (token)
does not see every doctor---a triage nurse (the router) assesses
symptoms and sends them to the relevant specialists. The hospital has
enormous total expertise (parameters), but each patient only uses a
small fraction. This is vastly more efficient than making every doctor
examine every patient.
\end{analogy}

% ----------------------------------------------------------------------------
\subsection{Where the Floating Point Lives}
\label{sec:where-fp-lives}
% ----------------------------------------------------------------------------

Now that we understand the full transformer architecture, we can
precisely identify where floating-point arithmetic is required and
where it might be eliminated. \Cref{tab:fp-map} provides this
breakdown, with cross-references to the figures and sections where
each component was explained.

\begin{table}[htbp]
\centering
\small
\caption{Floating-point dependency map of a transformer. The
  ``Where?'' column links each component to the figure or section
  where it was introduced. Recent research
  (I-BERT~\cite{kim2021ibert}, I-ViT~\cite{li2022ivit},
  I-LLM~\cite{hu2024illm}) has demonstrated integer-only
  approximations for components previously thought to require
  floating point.}
\label{tab:fp-map}
\begin{tabular}{@{}p{2.6cm}p{1.8cm}p{2.5cm}p{5.2cm}@{}}
  \toprule
  \textbf{Component} & \textbf{FP\,req'd?} & \textbf{Where?} & \textbf{Notes} \\
  \midrule
  Embedding lookup
    & No
    & \cref{sec:embeddings}; bottom of Fig.~\ref{fig:full-llm}
    & Pure table lookup (integer index $\to$ vector). \\[4pt]
  QKV projection
    & Replaceable
    & \cref{sec:self-attention}, Eq.~\ref{eq:qkv}; step~1 in the attention pipeline
    & Matrix multiply; BitNet uses ternary weights. \\[4pt]
  $\bm{Q}\bm{K}^\top$ dot product
    & Replaceable
    & \cref{sec:self-attention}, Eq.~\ref{eq:attention}; Fig.~\ref{fig:attention-scores}
    & Integer dot product with ternary weights works. \\[4pt]
  Softmax
    & \textbf{Repl.}
    & Eq.~\ref{eq:softmax}; Fig.~\ref{fig:softmax}
    & Integer approx.\ via polynomial~\cite{kim2021ibert} or
      bit-shift~\cite{li2022ivit}. Softmax-free alternatives
      also viable. See \cref{sec:softmax-problem}. \\[4pt]
  Value aggregation
    & Replaceable
    & Step~5 in \cref{sec:self-attention}
    & Weighted sum; integer weights from approx.\ softmax suffice. \\[4pt]
  Output projection
    & Replaceable
    & $\bm{W}_O$ in \cref{sec:self-attention}; step~4 of multi-head attention
    & Matrix multiply; can use ternary weights. \\[4pt]
  Layer norm
    & \textbf{Repl.}
    & \cref{sec:residual-layernorm}; Fig.~\ref{fig:full-llm}
    & Integer norm via I-BERT~\cite{kim2021ibert},
      I-ViT~\cite{li2022ivit}, I-LLM~\cite{hu2024illm}.
      See \cref{sec:layernorm-problem}. \\[4pt]
  MLP weights
    & Replaceable
    & \cref{sec:ffn}; $\bm{W}_1$, $\bm{W}_2$ in
      Figs.~\ref{fig:ffn},~\ref{fig:ffn-neurons}
    & BitNet uses ternary $\{-1,0,+1\}$. \\[4pt]
  MLP activation
    & Replaceable
    & $\sigma$ in \cref{sec:ffn}; Fig.~\ref{fig:activations}
    & LUT-compiled or bit-shift approx.\
      (ShiftGELU~\cite{li2022ivit}). \\[4pt]
  MoE gating
    & Replaceable
    & \cref{sec:moe}; Fig.~\ref{fig:moe}
    & Small classification; softmax over few experts. \\
  \bottomrule
\end{tabular}
\end{table}

\begin{keyinsight}
A striking finding from recent research: \emph{every} component in
a transformer has a demonstrated integer-only or bitwise
approximation. The attention softmax can be replaced by polynomial
approximations~\cite{kim2021ibert}, bit-shift
operations~\cite{li2022ivit}, or eliminated entirely in favour of
softmax-free attention mechanisms. Layer normalisation can be
implemented with integer arithmetic and bit-shifts. The question is
no longer ``can we build an integer-only transformer?'' but
``how much accuracy do the combined approximations cost, and can
we close that gap?'' \Cref{sec:scaling-llms} explores this in
detail.
\end{keyinsight}

% ----------------------------------------------------------------------------
% Full-page landscape diagram: the complete LLM architecture
% ----------------------------------------------------------------------------

\begin{landscape}
\begin{figure}[p]
\centering
\begin{tikzpicture}[>=Stealth, x=1cm, y=1cm,
  % --- Shared styles ---
  comp/.style={draw, rounded corners=3pt, minimum height=0.7cm,
    font=\scriptsize\sffamily, align=center, inner sep=4pt},
  smallcomp/.style={draw, rounded corners=2pt, minimum height=0.55cm,
    font=\tiny\sffamily, align=center, inner sep=3pt},
  arr/.style={-{Stealth[length=4pt]}, thick},
  fatarr/.style={-{Stealth[length=5pt]}, very thick},
  skiparr/.style={thick, dashed, bitgray},
  plus/.style={circle, draw, minimum size=0.4cm, font=\tiny,
    fill=bitgreen!10, inner sep=0pt},
  neuron/.style={draw, circle, minimum size=0.25cm, inner sep=0pt,
    fill=#1!15},
  brace/.style={decorate, decoration={brace, amplitude=5pt, mirror}},
  bracel/.style={decorate, decoration={brace, amplitude=5pt}},
  lbl/.style={font=\tiny\sffamily},
  heading/.style={font=\scriptsize\sffamily\bfseries}
]

% ============================================================================
% COLUMN 1: TOKENISER + EMBEDDING (x = 0..4)
% ============================================================================

% --- Tokeniser ---
\node[comp, fill=bitgray!8, draw=bitgray, minimum width=3.6cm] (tok)
  at (2, 0) {Tokeniser};
\node[lbl, below=1pt of tok, bitgray] {text $\to$ token IDs};

% BPE hint
\node[lbl, bitgray, text width=3.4cm, align=center] at (2, -1.2)
  {``The cat sat'' $\to$\\{}[464, 3857, 3290]};
\node[lbl, bitgray, text width=3.4cm, align=center] at (2, -2.0)
  {\textit{BPE vocabulary:}\\merge frequent byte pairs};

% --- Embedding table ---
\node[comp, fill=bitpurple!8, draw=bitpurple, minimum width=3.6cm] (emb)
  at (2, 2) {Embedding table $\bm{E}$};
\node[lbl, below right=1pt and -1.3cm of emb, bitpurple]
  {$V \!\times\! d$ lookup};
\draw[fatarr] (tok) -- (emb);

% Show lookup table sketch
\draw[bitpurple!40, thick] (0, 3.0) rectangle (4, 4.6);
\foreach \i/\lbl in {0/ID 0, 1/ID 1, 2/\dots, 3/ID $V\!-\!1$} {
  \pgfmathsetmacro{\yy}{3.15 + \i * 0.35}
  \node[lbl, bitpurple!70, anchor=west] at (0.15, \yy) {\lbl};
  \ifnum\i<3
    \draw[bitpurple!20] (0, {\yy + 0.27}) -- (4, {\yy + 0.27});
  \fi
}
\node[lbl, bitpurple, anchor=west] at (1.3, 3.15) {$\leftarrow d$ dims$\rightarrow$};
\draw[arr, bitpurple] (emb.north) -- (2, 3.0);

% Arrow up from embedding table to transformer blocks
\draw[fatarr] (2, 4.6) -- (2, 5.4);
\node[lbl, bitgray, anchor=west] at (2.2, 5.0) {$n \times d$ matrix};

% ============================================================================
% COLUMN 2: EXPANDED TRANSFORMER BLOCK 1 (x = 5..21)
% ============================================================================

% --- Block 1 outer frame ---
\draw[bitblue!50, thick, rounded corners=6pt, fill=bitblue!3]
  (4.5, -0.5) rectangle (21.5, 14.5);
\node[heading, bitblue, anchor=north west] at (4.7, 14.35)
  {Transformer Block 1 (expanded)};

% --- Input to block ---
\coordinate (blkin) at (13, -0.5);

% ========== SELF-ATTENTION SUB-LAYER (bottom half) ==========
\node[heading, bitblue!80, anchor=west] at (5.0, -0.1) {Self-Attention};

% Input tokens hint
\node[lbl, bitgray] at (7.5, 0.3) {$\bm{X}$ ($n \times d$)};

% Q K V projections
\node[smallcomp, fill=bitblue!12, draw=bitblue, minimum width=2cm]
  (wq) at (7, 1.5) {$\times\,\bm{W}_Q$};
\node[smallcomp, fill=bitgreen!12, draw=bitgreen, minimum width=2cm]
  (wk) at (10.5, 1.5) {$\times\,\bm{W}_K$};
\node[smallcomp, fill=bitorange!12, draw=bitorange, minimum width=2cm]
  (wv) at (14, 1.5) {$\times\,\bm{W}_V$};

% Arrows from X to Q,K,V
\foreach \nd in {wq, wk, wv} {
  \draw[arr] (7.5, 0.55) -- (\nd.south);
}
\node[lbl, bitblue] at (7, 2.05) {$\bm{Q}$};
\node[lbl, bitgreen] at (10.5, 2.05) {$\bm{K}$};
\node[lbl, bitorange] at (14, 2.05) {$\bm{V}$};

% Multi-head split hint
\draw[bracel, bitgray, thick] (5.8, 2.2) -- (15.2, 2.2);
\node[lbl, bitgray, above=3pt] at (10.5, 2.2)
  {split into $h$ heads (each $d_k = d/h$)};

% --- Dot product QK^T ---
\node[smallcomp, fill=bitblue!8, draw=bitblue, minimum width=2.4cm]
  (qk) at (8.75, 3.2) {$\bm{Q}\bm{K}^\top$};
\draw[arr, bitblue] (7, 2.1) -- (7, 2.8) -| (qk.south west);
\draw[arr, bitgreen] (10.5, 2.1) -- (10.5, 2.8) -| (qk.south east);

% --- Scale ---
\node[smallcomp, fill=bitgray!8, draw=bitgray, minimum width=2.4cm]
  (scale) at (8.75, 4.2) {$\div\,\sqrt{d_k}$};
\draw[arr] (qk) -- (scale);

% --- Softmax (with small curve) ---
\node[smallcomp, fill=bitorange!8, draw=bitorange, minimum width=2.4cm]
  (sfm) at (8.75, 5.2) {softmax};
\draw[arr] (scale) -- (sfm);

% Tiny softmax curve
\begin{scope}[shift={(10.5, 5.05)}, scale=0.3]
  \draw[bitorange, thick, smooth] plot[domain=-2:2, samples=30]
    ({\x}, {1/(1+exp(-3*\x))});
  \draw[gray, thin] (-2, 0) -- (2, 0);
\end{scope}

% --- Value aggregation ---
\node[smallcomp, fill=bitorange!8, draw=bitorange, minimum width=2.4cm]
  (agg) at (11.75, 5.2) {$\times\,\bm{V}$};
\draw[arr] (sfm) -- (agg);
\draw[arr, bitorange] (14, 2.1) -- (14, 4.8) -| (agg.south);

% --- Multi-head dots ---
\node[font=\scriptsize, bitgray] at (15.5, 3.6) {$\times h$};
\node[font=\scriptsize, bitgray] at (15.5, 3.0) {heads};
\draw[bitgray, thick, dotted] (14.8, 2.5) -- (14.8, 5.6);
\node[font=\large, bitgray] at (16.2, 4.2) {$\cdots$};

% Concat
\node[smallcomp, fill=bitgray!8, draw=bitgray, minimum width=2.0cm]
  (concat) at (11.75, 6.2) {concat};
\draw[arr] (agg) -- (concat);

% Output projection W_O
\node[smallcomp, fill=bitblue!10, draw=bitblue, minimum width=2.4cm]
  (wo) at (11.75, 7.1) {$\times\,\bm{W}_O$};
\draw[arr] (concat) -- (wo);

% --- Residual + LayerNorm 1 ---
\node[plus] (add1) at (11.75, 8.0) {$+$};
\draw[arr] (wo) -- (add1);

% Skip connection from input
\draw[skiparr] (5.5, 0.3) -- (5.5, 8.0) -- (add1.west);
\node[lbl, bitgray, rotate=90, anchor=south] at (5.2, 4.2)
  {residual (skip)};

\node[smallcomp, fill=bitgray!10, draw=bitgray, minimum width=2.8cm]
  (ln1) at (11.75, 8.9) {Layer Norm};
\draw[arr] (add1) -- (ln1);

% ========== MLP SUB-LAYER (top half) ==========
\node[heading, bitorange!80, anchor=west] at (5.0, 9.5)
  {Feed-Forward Network (MLP)};

% W1 expand
\node[smallcomp, fill=bitorange!12, draw=bitorange, minimum width=2.4cm]
  (mlpw1) at (9, 10.2) {$\times\,\bm{W}_1$ \textit{expand}};
\draw[arr] (ln1.north) -| (mlpw1.south);
\node[lbl, bitorange] at (9, 10.7) {$d \to 4d$};

% Activation
\node[smallcomp, fill=bitgreen!12, draw=bitgreen, minimum width=2.0cm]
  (act) at (13, 10.2) {$\sigma$ (GELU)};
\draw[arr] (mlpw1) -- (act);

% Tiny GELU curve
\begin{scope}[shift={(14.7, 10.05)}, scale=0.28]
  \draw[bitgreen, thick, smooth] plot[domain=-2.5:2.5, samples=30]
    ({\x}, {\x/(1+exp(-1.702*\x))});
  \draw[gray, thin] (-2.5, 0) -- (2.5, 0);
\end{scope}

% W2 compress
\node[smallcomp, fill=bitorange!12, draw=bitorange, minimum width=2.4cm]
  (mlpw2) at (17, 10.2) {$\times\,\bm{W}_2$ \textit{compress}};
\draw[arr] (act) -- (mlpw2);
\node[lbl, bitorange] at (17, 10.7) {$4d \to d$};

% Neuron dots (abstract MLP representation)
\foreach \i in {1,...,3} {
  \node[neuron=bitorange] at ({8.2 + \i*0.35}, 11.3) {};
}
\node[font=\tiny, bitorange] at (10.2, 11.3) {$\cdots$};
\foreach \i in {1,...,6} {
  \node[neuron=bitorange] at ({11.0 + \i*0.35}, 11.3) {};
}
\node[font=\tiny, bitorange] at (14.3, 11.3) {$\cdots$};
\foreach \i in {1,...,3} {
  \node[neuron=bitorange] at ({15.0 + \i*0.35}, 11.3) {};
}
\node[lbl, bitorange] at (12.5, 11.8) {$d$ $\to$ $4d$ $\to$ $d$ neurons};

% --- Residual + LayerNorm 2 ---
\node[plus] (add2) at (17, 12.0) {$+$};
\draw[arr] (mlpw2.north) -- (17, 11.5) -- (add2.south);

% Skip connection from LN1
\draw[skiparr] (5.8, 8.9) -- (5.8, 12.0) -- (add2.west);
\node[lbl, bitgray, rotate=90, anchor=south] at (5.5, 10.5)
  {residual};

\node[smallcomp, fill=bitgray!10, draw=bitgray, minimum width=2.8cm]
  (ln2) at (17, 13.0) {Layer Norm};
\draw[arr] (add2) -- (ln2);

% Output of block 1
\draw[fatarr] (ln2.north) -- (17, 14.5);

% ============================================================================
% COLUMN 3: STACKED BLOCKS 2..N + OUTPUT HEAD (x = 22..27)
% ============================================================================

% Arrow from block 1 to block stack
\draw[fatarr] (17, 14.5) -- (17, 15.2) -| (24, 14.5);

% Block 2
\node[comp, fill=bitblue!8, draw=bitblue, minimum width=3.8cm]
  (b2) at (24, 13.8) {Transformer Block 2};
\draw[arr] (24, 14.5) -- (b2.north);

% Dots
\node[font=\sffamily, bitgray] (bdots) at (24, 12.6) {$\vdots$};
\draw[arr] (b2.south) -- (bdots.north);

% Block N
\node[comp, fill=bitblue!8, draw=bitblue, minimum width=3.8cm]
  (bn) at (24, 11.4) {Transformer Block $N$};
\draw[arr] (bdots.south) -- (bn.north);

% Brace for N blocks
\draw[brace, thick, bitblue] (26.2, 14.2) -- (26.2, 11.1);
\node[lbl, bitblue, anchor=west, text width=1.5cm] at (26.6, 12.65)
  {$N$ layers\\(32--126)};

% --- Final layer norm ---
\node[comp, fill=bitgray!8, draw=bitgray, minimum width=3.8cm]
  (finln) at (24, 9.8) {Final Layer Norm};
\draw[fatarr] (bn.south) -- (finln.north);

% --- Output head ---
\node[heading, bitgreen!80] at (24, 9.0) {Output Head};

% W_vocab projection
\node[comp, fill=bitgreen!10, draw=bitgreen, minimum width=3.8cm]
  (wvoc) at (24, 8.0) {$\times\,\bm{W}_{\text{vocab}}$};
\draw[arr] (finln.south) -- (wvoc.north);
\node[lbl, bitgreen] at (24, 7.45) {$d \to V$ projection};

% Logits
\node[comp, fill=bitgreen!6, draw=bitgreen, minimum width=3.8cm]
  (logits) at (24, 6.4) {logits ($V$ raw scores)};
\draw[arr] (wvoc.south) -- (logits.north);

% Softmax
\node[comp, fill=bitorange!10, draw=bitorange, minimum width=3.8cm]
  (outsm) at (24, 5.0) {softmax};
\draw[arr] (logits.south) -- (outsm.north);

% Tiny softmax curve for output
\begin{scope}[shift={(26.4, 4.85)}, scale=0.3]
  \draw[bitorange, thick, smooth] plot[domain=-2:2, samples=30]
    ({\x}, {1/(1+exp(-3*\x))});
  \draw[gray, thin] (-2, 0) -- (2, 0);
\end{scope}

% Probabilities
\node[comp, fill=bitorange!6, draw=bitorange, minimum width=3.8cm]
  (probs) at (24, 3.6) {probabilities ($\sum\!=\!1$)};
\draw[arr] (outsm.south) -- (probs.north);

% Probability bar chart (mini)
\begin{scope}[shift={(22.1, 2.0)}]
  \foreach \i/\h/\c in {0/0.6/bitorange, 1/0.22/bitorange!60,
    2/0.08/bitorange!40, 3/0.04/bitorange!25, 4/0.02/bitorange!15} {
    \fill[\c] ({\i*0.42}, 0) rectangle ({\i*0.42+0.32}, {\h*1.5});
  }
  \node[font=\tiny, bitorange] at (2.5, 0.5) {$\cdots$};
  \node[lbl, bitorange] at (1.5, -0.3) {$V$ tokens};
\end{scope}

% Sampled token
\node[comp, fill=bitgreen!15, draw=bitgreen, minimum width=3.8cm]
  (sample) at (24, 0.8) {sample $\to$ next token};
\draw[arr] (probs.south) -- (24, 2.9) -- (24, 1.15);

% ============================================================================
% AUTOREGRESSIVE LOOP ARROW
% ============================================================================

% Arrow from sampled token looping back to tokeniser input
\draw[fatarr, bitgreen!70!black, rounded corners=8pt]
  (24, 0.45) -- (24, -1.6) -- (2, -1.6) -- (2, -0.35);
\node[lbl, bitgreen!70!black, fill=white, inner sep=2pt] at (13, -1.6)
  {autoregressive loop: append token, repeat};

% ============================================================================
% CONNECTING ARROWS between columns
% ============================================================================

% Embedding output to Block 1 input
\draw[fatarr] (2, 5.4) -- (2, 7.0) -- (7.5, 7.0) -- (7.5, 0.3);
\node[lbl, bitgray, fill=white, inner sep=1pt] at (4.8, 7.2)
  {+ positional encoding};

% ============================================================================
% TITLE AND ANNOTATIONS
% ============================================================================

\node[font=\small\sffamily\bfseries, bitblue!80!black, anchor=north west]
  at (0, 15.3) {The Complete LLM Architecture: End-to-End Pipeline};

% Dimension annotations
\node[lbl, bitgray, text width=3cm, align=center] at (2, 5.8)
  {$n$ tokens\\$d$ dims each};

\end{tikzpicture}
\caption[The complete LLM architecture in a single view.]{%
  The complete LLM architecture in a single view, integrating all
  components from this section. \textbf{Left:} the tokeniser converts
  text to integer IDs, which index into the embedding table $\bm{E}$
  ($V \times d$). \textbf{Centre:} Transformer Block~1 is expanded to
  show the self-attention sub-layer (Q/K/V projections, scaled dot
  product, softmax, value aggregation, output projection $\bm{W}_O$,
  with $h$ parallel heads) and the MLP sub-layer ($\bm{W}_1$ expand,
  GELU activation, $\bm{W}_2$ compress), each with residual
  connections and layer normalisation. \textbf{Right:} Blocks 2
  through $N$ are stacked compactly, followed by the final layer norm,
  output head ($\bm{W}_{\text{vocab}}$ projection $\to$ logits $\to$
  softmax $\to$ probabilities $\to$ sampled token), and the
  autoregressive loop that feeds each generated token back as input.
  Cross-references: attention mechanism
  (\cref{fig:attention-scores,fig:multi-head}), MLP block
  (\cref{fig:ffn,fig:ffn-neurons}), residual connections
  (\cref{fig:residual}), output head
  (\cref{fig:autoregressive}), component stack
  (\cref{fig:full-llm}).}
\label{fig:llm-full-picture}
\end{figure}
\end{landscape}
