% ============================================================================
\section{Scaling Bitwise Computation to Large Language Models}
\label{sec:scaling-llms}
% ============================================================================

Armed with an understanding of the transformer architecture, we can
now revisit the question from \cref{sec:hybrid}: can the four bitwise
paradigms be combined in a way that scales to LLM-sized models?

The previous section showed that only two components---the attention
softmax and layer normalisation---fundamentally require floating-point
arithmetic. This section proposes a combination strategy that assigns
each paradigm to the transformer component it is best suited for.

% ----------------------------------------------------------------------------
\subsection{Why No Single Paradigm Scales}
\label{sec:why-no-single}
% ----------------------------------------------------------------------------

Let us first be clear about why each paradigm \emph{alone} fails at
LLM scale:

\begin{description}[leftmargin=1em]
  \item[Binary Neural Networks] can binarise weights and activations,
    but the BNN approach assumes a fixed network topology. It does not
    address the attention mechanism, which is the defining innovation
    of transformers. BitNet b1.58~\cite{ma2024bitnet} succeeds
    precisely because it only binarises the \emph{weights}---not the
    architecture. The later 2B4T technical report provides an open,
    natively trained 2B model and CPU inference implementation, but it
    remains a ternary-weight rather than fully bitwise
    transformer~\cite{ma2025bitnet2b4t}.

  \item[Tsetlin Machines] learn propositional clauses over Boolean
    features. Language tokens are categorical with distributional
    semantics---booleanising 4{,}096-dimensional embeddings would
    destroy the fine-grained similarity structure that makes language
    modelling work. Additionally, Tsetlin Machines have no native
    mechanism for the pairwise token interactions that attention
    provides.

  \item[Logic Gate Networks] with 2-input gates would require
    astronomical circuit depth to approximate the $n \times n$
    attention computation over thousands of tokens. The softmax
    operation is inherently continuous. However, LGNs could work at
    the \emph{component level}---replacing individual MLP blocks
    rather than the entire transformer.

  \item[KAN / LUT compilation] faces an exponential scaling wall:
    lookup tables grow exponentially with input dimensionality.
    A function of 4{,}096-dimensional embeddings cannot be stored
    as a table. But a \emph{univariate} activation function (1D input)
    can easily be stored as a 256-entry table, which is exactly where
    LUT compilation shines.
\end{description}

\begin{keyinsight}
The insight is not ``which paradigm replaces the transformer?'' but
rather ``which paradigm can serve as a component \emph{inside} a
transformer?'' The transformer architecture is non-negotiable for
language. The four paradigms become optimisation techniques within
that structure, not replacements for it.
\end{keyinsight}

% ----------------------------------------------------------------------------
\subsection{A Combined Architecture: Bitwise Components in a Transformer Shell}
\label{sec:combined-architecture}
% ----------------------------------------------------------------------------

We propose mapping each paradigm to the transformer component where
its strengths are most relevant:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  comp/.style={draw, rounded corners=4pt, minimum width=5cm,
    minimum height=1.0cm, font=\small\sffamily, align=center,
    line width=0.8pt},
  paradigm/.style={font=\scriptsize\sffamily\bfseries, rounded corners=2pt,
    draw, inner sep=3pt, fill=#1!15, draw=#1!60},
  arr/.style={-{Stealth[length=5pt]}, thick},
  note/.style={font=\scriptsize\sffamily, text width=5cm}
]
  % Components stack (with section/figure references)
  \node[comp, fill=bitgray!5] (emb) at (0, 0)
    {Embedding lookup\\[-2pt]{\tiny\S\,9.3, Fig.~\ref{fig:full-llm} bottom}};
  \node[comp, fill=bitblue!8] (qkv) at (0, 1.5)
    {QKV projection\\[-2pt]{\tiny\S\,9.5, Eq.~\ref{eq:qkv}}};
  \node[comp, fill=bitpurple!8, draw=bitpurple!40] (attn) at (0, 3)
    {Attention softmax\\[-2pt]{\tiny\S\,9.5, Figs.~\ref{fig:softmax},~\ref{fig:attention-scores}}};
  \node[comp, fill=bitblue!8] (oproj) at (0, 4.5)
    {Output projection\\[-2pt]{\tiny\S\,9.5.4, $\bm{W}_O$ in multi-head}};
  \node[comp, fill=bitpurple!8, draw=bitpurple!40] (ln) at (0, 6)
    {Layer normalisation\\[-2pt]{\tiny\S\,9.7.2, Fig.~\ref{fig:full-llm}}};
  \node[comp, fill=bitorange!8] (mlp) at (0, 7.5)
    {MLP / feed-forward\\[-2pt]{\tiny\S\,9.6, Figs.~\ref{fig:ffn},~\ref{fig:ffn-neurons}}};
  \node[comp, fill=bitgreen!8] (gate) at (0, 9)
    {MoE router (if used)\\[-2pt]{\tiny\S\,9.10, Fig.~\ref{fig:moe}}};

  % Paradigm labels
  \node[paradigm=bitgray, right] at (3.5, 0)
    {Integer table lookup (no paradigm needed)};
  \node[paradigm=bitblue, right] at (3.5, 1.5)
    {BitNet: ternary $\{-1, 0, +1\}$ weights};
  \node[paradigm=bitpurple, right] at (3.5, 3)
    {Shiftmax / Softpick / linear attention};
  \node[paradigm=bitblue, right] at (3.5, 4.5)
    {BitNet: ternary weights};
  \node[paradigm=bitpurple, right] at (3.5, 6)
    {Bit-shift RMSNorm (I-ViT / I-LLM)};
  \node[paradigm=bitorange, right] at (3.5, 7.5)
    {LGN inference + LUT activations};
  \node[paradigm=bitgreen, right] at (3.5, 9)
    {Tsetlin Machine gating};

  % Arrows
  \foreach \a/\b in {emb/qkv, qkv/attn, attn/oproj, oproj/ln, ln/mlp, mlp/gate} {
    \draw[arr] (\a) -- (\b);
  }

  % Legend
  \draw[thick, bitblue!40, fill=bitblue!8, rounded corners=2pt]
    (-3.5, -1.5) rectangle (-1.5, -1.0);
  \node[font=\scriptsize\sffamily, right] at (-1.3, -1.25)
    {= integer/ternary arithmetic (proven)};
  \draw[thick, bitpurple!40, fill=bitpurple!8, rounded corners=2pt]
    (-3.5, -2.2) rectangle (-1.5, -1.7);
  \node[font=\scriptsize\sffamily, right] at (-1.3, -1.95)
    {= integer via bit-shifts (demonstrated)};
\end{tikzpicture}
\caption{Proposed mapping of bitwise paradigms to transformer
  components. Green and blue components can be made (near-)bitwise
  today. Red components remain open problems requiring floating-point
  arithmetic or novel replacements.}
\label{fig:combined-architecture}
\end{figure}

\subsubsection{Component 1: BitNet Ternary Weights (BNN Paradigm)}

The weight matrices $\bm{W}_Q$, $\bm{W}_K$, $\bm{W}_V$,
$\bm{W}_O$, $\bm{W}_1$, $\bm{W}_2$ are constrained to ternary
values $\{-1, 0, +1\}$, following the BitNet b1.58
approach~\cite{ma2024bitnet}.

\textbf{What this buys:}
\begin{itemize}[leftmargin=2em]
  \item Weight storage drops from 16~bits to 1.58~bits per parameter
    (a 10$\times$ reduction).
  \item Matrix multiplication becomes addition and subtraction (no
    multiplier circuits needed).
  \item Proven at 3B+ parameter scale with competitive accuracy.
\end{itemize}

\textbf{What remains floating-point:} The activations (intermediate
values flowing between layers) remain in higher precision (typically
8-bit integers in BitNet). The ternary constraint applies only to the
stored weights.

\subsubsection{Component 2: LUT-Compiled Activations (KAN Paradigm)}

The activation functions in the MLP blocks (GELU, SiLU, etc.)
are fixed mathematical functions applied element-wise. Instead of
computing them with floating-point arithmetic, we can:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Train} with learnable activation functions
    (KAN-style B-splines or learnable piecewise-linear functions).
  \item \textbf{Compile} the learned functions into lookup tables:
    a 256-entry table maps each possible 8-bit input to a
    precomputed output.
\end{enumerate}

\textbf{What this buys:}
\begin{itemize}[leftmargin=2em]
  \item Activation computation becomes a single table lookup---pure
    memory access, no arithmetic.
  \item Each layer can learn its own optimal activation shape, rather
    than using one fixed function everywhere.
  \item The LUT fits in L1 cache (256 entries $\times$ 1~byte = 256~bytes).
\end{itemize}

\textbf{Why KANs do not replace the full MLP:} As discussed in
\cref{sec:ffn}, the MLP block has $d \times d_\text{ff}$ connections.
Replacing each with a full B-spline would create an explosion in
parameters and training cost. But replacing only the
\emph{per-neuron activation function} with a learned spline and then
compiling to a LUT is lightweight and practical.

\subsubsection{Component 3: Logic Gate Network MLP (LGN Paradigm)}

The MLP blocks account for roughly two-thirds of the parameters and
computation in a transformer. After training with ternary weights and
LUT activations, the MLP can potentially be \emph{further compiled}
into a logic gate network for inference:

\begin{enumerate}[leftmargin=2em]
  \item Train the MLP with standard methods (ternary weights + LUT
    activations).
  \item After training, \emph{compile} the trained MLP into an
    equivalent Boolean circuit, where the quantised weights and
    activations are mapped to logic gate configurations.
  \item The inference-time MLP becomes a fixed logic circuit with no
    arithmetic operations.
\end{enumerate}

\textbf{What this buys:}
\begin{itemize}[leftmargin=2em]
  \item Inference-time MLP is pure combinational logic (AND/OR/XOR
    gates)---maximum hardware efficiency.
  \item Replaces selected operations with Boolean primitives.
  \item May reduce model storage when a sparse gate graph replaces a
    dense projection.
\end{itemize}

\textbf{What is unproven:} Recent LGNs have reached CIFAR-100 and
connection-optimised MNIST results~\cite{ruttgers2025light,mommen2026fullytrainable},
but not transformer MLP scale. A useful CPU implementation must also show
that its sparse graph traversal beats a compact SIMD integer kernel; fewer
Boolean operations alone is not enough.

\subsubsection{Component 4: Tsetlin Machine Router (TM Paradigm)}

In a Mixture-of-Experts transformer, the gating network decides which
expert processes each token. This is a small classification problem:
given a token's hidden state, select the top-$k$ experts from a pool
of $E$ options.

Tsetlin Machines are a natural fit here because:
\begin{itemize}[leftmargin=2em]
  \item The routing decision is a low-dimensional classification
    (typically $E = 8$ to 64 experts).
  \item The input can be booleanised (e.g., by thresholding the
    hidden state dimensions) without catastrophic information loss,
    because the routing decision does not need the same fine-grained
    precision as the main computation.
  \item The learned clauses would be \emph{interpretable}: we could
    inspect \emph{why} a particular token was routed to a particular
    expert, expressed as human-readable Boolean rules.
\end{itemize}

\textbf{What this buys:}
\begin{itemize}[leftmargin=2em]
  \item Interpretable routing decisions (no black-box gating).
  \item Integer-only routing computation (no floating-point softmax
    in the gating network).
  \item Potential for faster routing than a learned linear layer.
\end{itemize}

\textbf{What is unproven:} Whether Tsetlin Machine routing can match
the load-balancing and accuracy of continuous gating functions at LLM
scale.

% ----------------------------------------------------------------------------
\subsection{The Softmax Problem}
\label{sec:softmax-problem}
% ----------------------------------------------------------------------------

The attention softmax (\cref{eq:softmax}) was initially seen as the
single most important obstacle to a fully bitwise transformer. Recent
research has dramatically changed this picture: multiple approaches
now exist to implement softmax---or replace it entirely---without
floating-point arithmetic.

\subsubsection{Why Softmax Was Thought to Require Floating Point}

Softmax converts a vector of raw scores $\bm{z} = [z_1, \ldots, z_n]$
into a probability distribution:

\[
  \text{softmax}(z_i) = \frac{e^{z_i}}{\sum_{j=1}^{n} e^{z_j}}
\]

Three properties make this seemingly incompatible with integer
arithmetic:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Exponentiation.} The function $e^x$ maps integers to
    irrational numbers, with unbounded dynamic range.

  \item \textbf{Normalisation.} Dividing by the sum ensures outputs
    sum to~1. Division is expensive and does not have a trivial
    bitwise equivalent.

  \item \textbf{Precision sensitivity.} Transformers produce
    occasional ``mega-activations''---outlier values orders of
    magnitude larger than their peers---which cause catastrophic
    quantisation errors when softmax is naively implemented in low
    precision.
\end{enumerate}

\subsubsection{Approach 1: Integer-Only Softmax Approximation}

A series of papers has demonstrated that softmax \emph{can} be
computed using only integer arithmetic and bit-shift operations:

\begin{description}[leftmargin=1em]
  \item[I-BERT~\cite{kim2021ibert}] (2021) was the first to achieve
    fully integer-only inference for a transformer (BERT). It
    approximates $e^x$ using a second-order polynomial evaluated
    entirely in INT32 arithmetic, achieving a maximum error of only
    $1.9 \times 10^{-3}$. Layer normalisation is handled with an
    integer square-root algorithm. The result: 2.4--4.0$\times$
    speedup on GPU with no accuracy loss.

  \item[I-ViT~\cite{li2022ivit}] (2022) introduced
    \concept{Shiftmax}---a softmax implementation that replaces
    the exponential with bit-shift approximations. Combined with
    \concept{ShiftGELU} (a bit-shift approximation of the GELU
    activation) and integer LayerNorm, the entire Vision Transformer
    runs on integer arithmetic (INT8 for linear operations, INT32
    for nonlinear operations) with no floating-point instructions.
    Result: 3.7--4.1$\times$ inference speedup with accuracy
    matching or exceeding the floating-point baseline.

  \item[I-LLM~\cite{hu2024illm}] (2024) extended integer-only
    inference to \emph{large language models} specifically. It
    introduces DI-ClippedSoftmax (a ``dynamic integer'' variant
    that handles the mega-activation problem) and DI-Exp (bit-shift
    exponentiation). All nonlinear operations---softmax, SwiGLU
    activation, layer normalisation---run on integer arithmetic and
    bit-shifting. At W6A6 (6-bit weights and activations), accuracy
    closely matches the floating-point baseline; at W4A4, it
    significantly outperforms other quantisation methods, though
    some gap to full precision remains.
\end{description}

\begin{engineernote}
The I-BERT $\to$ I-ViT $\to$ I-LLM progression shows a clear
trajectory: integer-only softmax went from BERT-scale (2021) to
Vision Transformers (2022) to LLMs (2024) in just three years.
The bit-shift approach from I-ViT is particularly relevant for
our bitwise agenda: \emph{bit-shifting is a bitwise operation}.
A transformer using Shiftmax computes attention without any
multiplication at all---only shifts and additions.
\end{engineernote}

\subsubsection{Approach 2: Softmax-Free Attention}

Rather than approximating softmax in integer arithmetic, a more
radical approach is to \emph{eliminate} it entirely:

\begin{description}[leftmargin=1em]
  \item[Linear attention~\cite{katharopoulos2020linear}.]
    Replace $\text{softmax}(\bm{Q}\bm{K}^\top)\bm{V}$
    with $\phi(\bm{Q})(\phi(\bm{K})^\top\bm{V})$, where $\phi$ is a
    feature map (e.g., $\text{elu}(x) + 1$). By changing the order
    of operations (parenthesisation), the $n \times n$ attention
    matrix is never explicitly computed. Complexity drops from
    $O(n^2)$ to $O(n)$. The TransNormerLLM~\cite{qin2023transnormerllm}
    showed that linear attention with gating mechanisms can
    \emph{outperform} softmax-attention LLMs in both accuracy and
    efficiency at up to 7B parameters.

  \item[ReLU attention~\cite{wortsman2023relu}.]
    Replace softmax with
    $\text{ReLU}(\bm{Q}\bm{K}^\top) / L$ (where $L$ is the
    sequence length). ReLU is a simple threshold: output the input
    if positive, zero otherwise. This is trivially implementable
    with bitwise operations (check the sign bit). On vision tasks,
    ReLU attention matches softmax scaling on
    ImageNet-21k~\cite{wortsman2023relu}. Whether this transfers
    to language modelling is an open question---the cited paper
    evaluates only vision transformers.

  \item[Softpick~\cite{fuadi2025softpick}] (2025) takes a
    targeted approach: it drops softmax's sum-to-one constraint
    while keeping the exponential, producing a ``rectified softmax.''
    This eliminates two pathological behaviours---attention
    sinks\footnote{Tokens forced to attend \emph{somewhere} even
    when no token is relevant, creating spurious attention to
    positional tokens.} and mega-activations---that plague
    quantisation. At GPTQ 2-bit quantisation, Softpick achieves
    dramatically better perplexity than standard softmax (which
    degrades catastrophically at this precision). It also produces
    highly sparse attention maps (95--99\% sparsity), substantially
    reducing computation. This is the most promising ``near-softmax''
    replacement for quantisation-friendly attention.

  \item[Polynomial attention~\cite{poly2024softmax}.]
    Recent work shows that softmax's effectiveness comes from
    implicit Frobenius norm regularisation, \emph{not} from
    producing a probability distribution. Polynomial functions
    can provide the same regularisation effect with far simpler
    arithmetic. Polynomials are straightforward to evaluate in
    integer arithmetic (only multiply and add).

  \item[Gated attention / state-space models.] Architectures like
    Mamba~\cite{gu2024mamba} and RWKV replace attention entirely
    with recurrent state-space mechanisms that use element-wise
    gating instead of softmax. These are inherently more amenable
    to integer implementation, as gating is a simple multiplication
    (or, with binary gates, a simple mask). Mamba has shown
    competitive results at 1.4B parameters, matching transformers
    of twice its size on some benchmarks.
\end{description}

\begin{keyinsight}
The softmax problem is \emph{solved} in the engineering sense---there
are multiple demonstrated paths:
\begin{enumerate}[leftmargin=1em]
  \item \textbf{Keep softmax, make it integer:} Shiftmax (I-ViT)
    and DI-ClippedSoftmax (I-LLM) compute attention using only
    bit-shifts and integer addition. No floating point needed.
  \item \textbf{Replace softmax with something friendlier:}
    Softpick, polynomial attention, or ReLU attention eliminate
    the exponential entirely.
  \item \textbf{Remove attention altogether:} Linear attention
    (TransNormerLLM) and state-space models (Mamba) avoid the
    $n \times n$ attention matrix, making the integer arithmetic
    even simpler.
\end{enumerate}
The remaining question is not ``can it be done?'' but ``which
combination of approach and architecture yields the best
accuracy-efficiency trade-off at scale?''
\end{keyinsight}

% ----------------------------------------------------------------------------
\subsection{The Layer Normalisation Problem}
\label{sec:layernorm-problem}
% ----------------------------------------------------------------------------

Layer normalisation (\cref{eq:layernorm}) was the second
suspected floating-point holdout. It requires computing the mean and
variance of a vector, then dividing each element by the standard
deviation. This problem turns out to be largely solved.

\begin{description}[leftmargin=1em]
  \item[RMSNorm] (used in Llama and most modern LLMs) drops the
    mean computation entirely:
    $\text{RMSNorm}(\bm{x}) = \bm{x} / \text{RMS}(\bm{x}) \cdot \bm{\gamma}$,
    where $\text{RMS}(\bm{x}) = \sqrt{\frac{1}{d}\sum_i x_i^2}$.
    This still requires a square root and division, but the
    computation is simpler than full LayerNorm.

  \item[Integer LayerNorm (I-BERT~\cite{kim2021ibert}).]
    The sum of squares $\sum_i x_i^2$ is an integer operation if
    $\bm{x}$ is integer. I-BERT demonstrated an integer square-root
    algorithm and integer division that implement full LayerNorm
    in INT32 arithmetic only. The same technique applies to the
    simpler RMSNorm.

  \item[Integer LayerNorm via bit-shifts (I-ViT~\cite{li2022ivit},
    I-LLM~\cite{hu2024illm}).] I-ViT's I-LayerNorm and
    I-LLM's DI-Norm implement normalisation using integer
    arithmetic with bit-shift-based square root approximation.
    I-LLM extends this to LLM-scale RMSNorm. The approach
    replaces expensive floating-point operations with iterative
    integer algorithms.

  \item[Pragmatic argument.] Even if normalisation stayed in
    floating point, it accounts for less than 1\% of total
    operations in a transformer. BitNet b1.58 retains RMSNorm
    in higher precision for this reason. But with I-ViT's
    approach, even this last fraction can be eliminated.
\end{description}

\begin{engineernote}
Layer normalisation is a \textbf{solved problem} for integer-only
inference. I-BERT, I-ViT, and I-LLM have each demonstrated integer
normalisation with no meaningful accuracy loss. The only remaining
question is whether integer normalisation during \emph{training}
(not just inference) causes instability---but for deployment of a
trained model, integer normalisation works today.
\end{engineernote}

% ----------------------------------------------------------------------------
\subsection{What BitNet Actually Binarises (and What It Does Not)}
\label{sec:bitnet-reality}
% ----------------------------------------------------------------------------

Before assessing the overall picture, it is important to understand
precisely what BitNet b1.58~\cite{ma2024bitnet} does and does not
binarise, since it is often described simply as ``1-bit LLM'':

\begin{table}[htbp]
\centering
\caption{What BitNet b1.58 binarises vs.\ what remains in higher
  precision. The ``1-bit'' label applies only to stored weight
  matrices, not to the full computation.}
\label{tab:bitnet-reality}
\begin{tabular}{@{}lll@{}}
  \toprule
  \textbf{Component} & \textbf{BitNet b1.58} & \textbf{Precision} \\
  \midrule
  Linear projection weights & Ternary $\{-1,0,+1\}$ & 1.58 bits \\
  Activations (inter-layer) & Quantised & 8-bit integer \\
  Attention softmax & Standard & FP16 \\
  RMSNorm & Standard & FP16 \\
  Scaling factors & Standard & FP16 \\
  Embedding layer & Standard & FP16 \\
  \bottomrule
\end{tabular}
\end{table}

Recent extensions push further: BitNet
a4.8~\cite{wang2024bitneta48} reduces activations to 4-bit with
hybrid sparsification (only 55\% of parameters activated), and
QuEST~\cite{panferov2025quest} demonstrates stable
quantisation-aware training at 1-bit for \emph{both} weights and
activations.

The gap between ``BitNet as published'' and ``fully integer-only
transformer'' is exactly the softmax and normalisation components---which,
as we have just seen, both have demonstrated integer solutions.

% ----------------------------------------------------------------------------
\subsection{The Realistic Picture}
\label{sec:realistic-picture}
% ----------------------------------------------------------------------------

\Cref{tab:bitwise-llm-summary} summarises the current state of each
component, updated to reflect the integer-only attention and
normalisation research.

\begin{table}[htbp]
\centering
\caption{Feasibility assessment for bitwise transformer components.
  ``Proven'' means demonstrated at LLM scale (billions of parameters);
  ``demonstrated'' means shown to work at meaningful scale with
  competitive accuracy; ``open'' means no convincing demonstration
  exists.}
\label{tab:bitwise-llm-summary}
\begin{tabular}{@{}p{3cm}lp{3cm}p{4.5cm}@{}}
  \toprule
  \textbf{Component} & \textbf{Paradigm} & \textbf{Status}
    & \textbf{Key challenge} \\
  \midrule
  Weight storage & BitNet (BNN) & Proven (2B+) & Ternary weights work; full model is not 1-bit. \\
  Matrix multiply & BitNet (BNN) & Proven & Addition/subtraction only. \\
  Activation fn. & KAN $\to$ LUT & Demonstrated & ShiftGELU~\cite{li2022ivit} or LUT compilation. \\
  MLP compilation & LGN & Open & CPU-friendly sparse traversal at transformer MLP scale. \\
  MoE routing & Tsetlin Machine & Open & Load balancing, accuracy matching. \\
  Attention softmax & Integer approx. & Demonstrated & Shiftmax~\cite{li2022ivit}, DI-ClippedSoftmax~\cite{hu2024illm} work. Integration with ternary weights untested. \\
  Softmax-free attn. & Linear / Mamba & Demonstrated & TransNormerLLM competitive at 7B scale~\cite{qin2023transnormerllm}. \\
  Layer norm & Bit-shift & Demonstrated & I-ViT / I-LLM show negligible accuracy loss. \\
  \bottomrule
\end{tabular}
\end{table}

\subsubsection{What Works Today (Separately)}

Each of the following has been independently demonstrated:
\begin{itemize}[leftmargin=2em]
  \item Ternary weights throughout (BitNet b1.58, demonstrated in an
    open 2B model~\cite{ma2025bitnet2b4t}).
  \item Integer-only softmax via bit-shifts (I-ViT Shiftmax,
    I-LLM DI-ClippedSoftmax).
  \item Integer-only layer normalisation via bit-shifts (I-ViT,
    I-LLM).
  \item Integer-only activation functions (ShiftGELU, or
    LUT compilation).
  \item Softmax-free attention at LLM scale (TransNormerLLM at 7B).
  \item 1-bit weights \emph{and} activations training (QuEST).
\end{itemize}

\emph{No one has combined all of these into a single end-to-end
integer-only LLM trained from scratch.} This integration is the
key open problem. Each individual component works, but compounding
approximation errors across 80+ layers with all components
simultaneously active is uncharted territory.

\subsubsection{The Fully Integer Transformer: Assembly Required}

The most promising configuration, based on existing research, would
combine:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Ternary weights} (BitNet b1.58) for all linear
    projections.
  \item \textbf{Shiftmax or Softpick} for attention (bit-shift or
    quantisation-friendly softmax).
  \item \textbf{Bit-shift RMSNorm} (from I-ViT/I-LLM) for
    normalisation.
  \item \textbf{LUT-compiled activations} (from KAN paradigm) or
    \textbf{ShiftGELU} (from I-ViT) for nonlinearities.
\end{enumerate}

This would produce a transformer where \emph{every operation} is
integer arithmetic, bit-shifting, or table lookup. No floating-point
unit would be needed at inference time.

\subsubsection{What Remains Genuinely Open}

\begin{itemize}[leftmargin=2em]
  \item \textbf{Integration at scale.} Combining ternary weights
    \emph{with} integer softmax \emph{with} bit-shift normalisation
    in a single model. The error from each approximation may compound.
  \item \textbf{Training.} Most integer-only results are
    inference-only. Training still typically requires higher
    precision. QuEST is the exception but has not been demonstrated
    beyond moderate scale.
  \item \textbf{LGN compilation of MLPs.} Logic Gate Networks now have
    CIFAR-100 and connection-optimised MNIST evidence, but transformer
    MLPs at
    7B scale have 67M parameters per layer---the numbers are
    comparable, but the computational structure differs fundamentally.
  \item \textbf{Tsetlin Machine MoE routing.} Tsetlin Machines
    work well for text classification but have not been applied to
    expert routing in a transformer.
\end{itemize}

% ----------------------------------------------------------------------------
\subsection{The Path Forward}
\label{sec:path-forward-llms}
% ----------------------------------------------------------------------------

Rather than attempting a fully bitwise LLM immediately, a staged
research programme would build confidence incrementally:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{MNIST hybrid} (current project): Validate the
    four-paradigm combination on a tractable problem.

  \item \textbf{BitNet + LUT activations}: Take an existing BitNet
    model and replace fixed activations with LUT-compiled learned
    activations. Measure accuracy impact. This is a minimal
    intervention with clear evaluation.

  \item \textbf{Softmax-free BitNet}: Combine ternary weights with
    a softmax-free attention mechanism (linear attention, ReLU
    attention, or Mamba-style gating). Evaluate on standard language
    benchmarks.

  \item \textbf{LGN MLP compilation}: Train a small transformer
    (125M parameters) with ternary weights, then attempt to compile
    the MLP blocks into logic gate networks. Measure inference
    speed and accuracy preservation.

  \item \textbf{TM routing for MoE}: Implement Tsetlin Machine
    routing in a MoE transformer and compare with standard softmax
    gating on load balance and accuracy.

  \item \textbf{Integration}: Combine the successful components
    into a single architecture and evaluate at increasing scale.
\end{enumerate}

Each step is independently publishable and valuable regardless of
whether the full vision materialises. The MNIST project we develop
in this article is step~1.
