% ============================================================================
% Bitwise Digits: Theory and Design of a Hybrid Logic-Based
% Neural Network for MNIST
% ============================================================================
\documentclass[11pt,a4paper,twoside,openright]{book}

% --- Shared preamble ---
\input{preamble}

% --- Document metadata ---
\title{%
  \vspace{-2cm}
  {\Huge\bfseries Bitwise Digits}\\[1em]
  {\Large Theory and Design of a Hybrid Logic-Based\\
   Neural Network for MNIST}\\[2em]
  {\large A Practical Guide for Engineers}
}
\author{Bernhard Gerlach}
\date{March 2026}

% ============================================================================
\begin{document}
% ============================================================================

\maketitle
\thispagestyle{empty}

% --- Front matter ---
\frontmatter

\chapter*{Preface}

This book describes the theory and design of a neural network that uses no
floating-point arithmetic at inference time. Instead of the
multiply-accumulate operations that dominate modern AI, our architecture
relies on Boolean logic, integer counting, and memory lookups---operations
that every CPU can perform cheaply and efficiently.

The architecture combines three paradigms:
\begin{itemize}
  \item \textbf{Tsetlin Machines} learn interpretable if--then rules using
    simple counting automata. They form the feature extraction layer.
  \item \textbf{Logic Gate Networks} learn which logic gates (AND, OR,
    XOR, \ldots) best combine those features. They form the composition
    layer.
  \item \textbf{Lookup Table compilation} converts any remaining learned
    functions into precomputed tables. It forms the deployment layer.
\end{itemize}

The target task is handwritten digit recognition on the MNIST dataset---a
deliberately simple benchmark chosen to validate the approach, not to chase
records. The implementation is in Rust.

\paragraph{Who this book is for.}
Engineers who are comfortable with university-level mathematics (linear
algebra, basic probability, logic) but who are not specialists in machine
learning or theoretical computer science. Every concept is introduced from
intuition and diagrams before it is formalised.

\paragraph{How to read this book.}
Part~I builds the foundations: Boolean algebra, supervised learning, and
data preparation. Parts~II--IV cover the three components of the hybrid
architecture in depth. Part~V assembles everything into a complete system
and analyses its properties. Each part is designed to be self-contained
enough that you can read the Tsetlin chapters (Part~II) and the Logic Gate
chapters (Part~III) in either order, provided you read both before Part~V.

\paragraph{Conventions.}
Throughout this book, coloured boxes highlight important ideas:
\begin{keyinsight}
Blue boxes contain key theoretical insights---the ideas that, once
understood, make everything else click.
\end{keyinsight}
\begin{engineernote}
Green boxes contain practical engineering observations---implementation
tips, performance implications, and hardware considerations.
\end{engineernote}
\begin{analogy}
Orange boxes contain analogies that map unfamiliar concepts to everyday
experience. They sacrifice precision for intuition.
\end{analogy}

\vfill
\noindent\textit{March 2026}

\tableofcontents

% --- Main matter ---
\mainmatter

% Part I: Foundations
\part{Foundations}
\label{part:foundations}

\input{chapters/ch01_problem}
\input{chapters/ch02_boolean}
\input{chapters/ch03_learning}
\input{chapters/ch04_mnist_prep}

% Part II: Tsetlin Machines
\part{Tsetlin Machines --- Learning Logic Without Gradients}
\label{part:tsetlin}

\input{chapters/ch05_automaton}
\input{chapters/ch06_clauses}
\input{chapters/ch07_tm_mnist}

% Part III: Logic Gate Networks
\part{Logic Gate Networks --- Learning the Circuit}
\label{part:lgn}

\input{chapters/ch08_diff_gates}
\input{chapters/ch09_lgn_training}

% Part IV: LUT Compilation
\part{LUT Compilation --- From Functions to Tables}
\label{part:lut}

\input{chapters/ch10_splines}
\input{chapters/ch11_lut}

% Part V: The Complete System
\part{The Complete Hybrid Architecture}
\label{part:hybrid}

\input{chapters/ch12_assembly}
\input{chapters/ch13_analysis}
\input{chapters/ch14_implementation}

% --- Back matter ---
\backmatter

% Appendices
\appendix
\part*{Appendices}
\addcontentsline{toc}{part}{Appendices}

\input{chapters/appA_notation}
\input{chapters/appB_gates}

% Bibliography
\printbibliography[heading=bibintoc]

\end{document}
