% ============================================================================
% Bitwise AI: A State-of-the-Art Overview
% Main document file
% ============================================================================
\documentclass[11pt,a4paper]{article}

% --- Core packages ---
\usepackage{fontspec}
\usepackage[english]{babel}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}

% --- Typography ---
\usepackage{microtype}
\setmainfont{Latin Modern Roman}
\setsansfont{Latin Modern Sans}
\setmonofont{Latin Modern Mono}

% --- Math ---
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{bm}

% --- Graphics and diagrams ---
\usepackage{tikz}
\usetikzlibrary{
  arrows.meta, positioning, calc, decorations.pathreplacing,
  shapes.geometric, shapes.misc, fit, backgrounds, patterns,
  matrix, chains, circuits.logic.US
}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% --- Tables and lists ---
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{enumitem}

% --- Colors ---
\usepackage{xcolor}
\definecolor{bitblue}{HTML}{2563EB}
\definecolor{bitgreen}{HTML}{16A34A}
\definecolor{bitorange}{HTML}{EA580C}
\definecolor{bitpurple}{HTML}{9333EA}
\definecolor{bitgray}{HTML}{6B7280}
\definecolor{lightbg}{HTML}{F8FAFC}

% --- Code and algorithms ---
\usepackage{algorithm}
\usepackage{algpseudocode}

% --- Hyperlinks and references ---
\usepackage[
  colorlinks=true,
  linkcolor=bitblue,
  citecolor=bitgreen,
  urlcolor=bitorange
]{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}

% --- Bibliography ---
\usepackage[
  backend=biber,
  style=numeric-comp,
  sorting=nyt,
  maxbibnames=99
]{biblatex}
\addbibresource{references.bib}

% --- Theorem environments ---
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]

% --- Custom commands ---
\newcommand{\bitop}[1]{\texttt{\textcolor{bitblue}{#1}}}
\newcommand{\concept}[1]{\textbf{\textcolor{bitpurple}{#1}}}
\newcommand{\xmark}{\textcolor{red}{\ding{55}}}

% --- Custom environments ---
\usepackage{tcolorbox}
\tcbuselibrary{skins, breakable}

\newtcolorbox{keyinsight}[1][]{
  enhanced, breakable,
  colback=bitblue!5, colframe=bitblue!70,
  fonttitle=\bfseries,
  title={Key Insight},
  #1
}

\newtcolorbox{engineernote}[1][]{
  enhanced, breakable,
  colback=bitgreen!5, colframe=bitgreen!70,
  fonttitle=\bfseries,
  title={Engineer's Note},
  #1
}

\newtcolorbox{analogy}[1][]{
  enhanced, breakable,
  colback=bitorange!5, colframe=bitorange!70,
  fonttitle=\bfseries,
  title={Analogy},
  #1
}

% --- Document metadata ---
\title{%
  \textbf{Beyond Multiply-Accumulate}\\[0.5em]
  \Large A Practical Guide to Bitwise Neural Computation\\[0.3em]
  \large An Overview of Four Paradigms for Hardware-Efficient AI
}
\author{Bernhard Gerlach}
\date{March 2026}

% ============================================================================
\begin{document}
% ============================================================================

\maketitle

\begin{abstract}
Modern neural networks are built almost entirely on one operation:
multiply a number by a weight and add the result to a running total.
This \emph{multiply-accumulate} (MAC) operation demands floating-point
hardware, consumes significant energy, and has led to an industry
dominated by GPU manufacturers. But what if neural networks could
be built from the simplest operations a processor can perform---AND,
OR, XOR, bit shifts, and table lookups?

This article surveys four paradigms that pursue this goal:
\textbf{Binary Neural Networks}, which constrain weights to $\{-1, +1\}$
and replace multiplication with XNOR gates;
\textbf{Tsetlin Machines}, which learn propositional logic clauses
through simple automata;
\textbf{Logic Gate Networks}, which learn the type of logic gates
end-to-end; and
\textbf{Kolmogorov-Arnold Networks with Lookup Table compilation},
which train complex learnable functions and then compile them
into memory tables for near-zero-cost inference.

Each paradigm is introduced from first principles, with diagrams
and analogies before equations. The article examines a fundamental
tension across these approaches: the trade-off between data
precision and operation type. Pure bitwise methods booleanise inputs
to single bits for maximum hardware efficiency; integer
quantisation methods keep multi-bit data but require costlier
arithmetic. The gap between these two camps---multi-bit data
processed with purely bitwise operations---remains an open research
question.

Beyond classification, the article provides a self-contained
introduction to transformer architectures and large language models
(LLMs), then evaluates whether bitwise paradigms can scale to
LLM-sized models. The analysis shows that every transformer
component---including the attention softmax and layer
normalisation---now has a demonstrated integer-only or bit-shift
implementation. A combined architecture is proposed: ternary
weights (BitNet), bit-shift attention, integer normalisation, and
LUT-compiled activations. The remaining challenge is integration
at scale, not the feasibility of individual components.

The article concludes with a hybrid architecture for MNIST digit
recognition as a proof-of-concept, targeting implementation
entirely in Rust with no GPU dependency.
\end{abstract}

\clearpage

\begin{small}
\noindent\textbf{Disclaimer: AI-Assisted Document.}\quad
This document was generated with substantial assistance from a large
language model (Claude, Anthropic). The human author directed the
research questions, architectural vision, and editorial decisions;
the AI assisted with literature research, technical writing, diagram
creation, and \LaTeX{} typesetting.

\smallskip\noindent\textit{Sources of information.}\quad
The content draws on three sources: (1)~the AI model's training data,
which includes academic publications up to its knowledge cutoff;
(2)~web searches conducted by AI agents during the writing process,
which retrieved paper abstracts and secondary summaries rather than
full-text PDFs; and (3)~the author's domain knowledge.

\smallskip\noindent\textit{Citation reliability.}\quad
All bibliography entries were verified against arXiv listings or
publisher metadata. Specific claims \emph{about} papers (accuracy
figures, speedup factors, architectural details) were subsequently
audited against the downloaded full-text PDFs where available, and
corrected where discrepancies were found. Claims about papers
without freely accessible PDFs (paywalled proceedings, books) carry
higher uncertainty. The reader is encouraged to consult the original
sources, available in the \texttt{research/papers/} directory of the
companion repository.
\end{small}

\clearpage

\tableofcontents
\clearpage

% --- Sections ---
\input{sections/01_introduction}
\input{sections/02_foundations}
\input{sections/03_binary_neural_networks}
\input{sections/04_tsetlin_machines}
\input{sections/05_logic_gate_networks}
\input{sections/06_kan_lut}
\input{sections/07_comparison}
\input{sections/08_hybrid_vision}
\input{sections/09_transformers_llms}
\input{sections/10_scaling_bitwise_llms}
\input{sections/11_conclusion}

% --- Abbreviations ---
\input{sections/99_abbreviations}

% --- Bibliography ---
\clearpage
\printbibliography[heading=bibintoc]

\end{document}
